---
title: "Observational data"
author: "Lucy D'Agostino McGowan"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    css: ["default", "css/style.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```


```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view"))
```

class: center middle main-title section-title-1

# Observational data

.class-info[

**Session 8**

.light[STA 379/679: Causal Inference <br>
Lucy D'Agostino McGowan
]

]

---


class: title title-1

# Causal inference

1. Ask an appropriate causal question

---

class: title title-1

# Causal inference

1. Ask an appropriate causal question
1. Draw your assumptions


---

class: title title-1

# Causal inference

1. Ask an appropriate causal question
1. Draw your assumptions
1. Organize your data


---

class: title title-1

# Causal inference

1. Ask an appropriate causal question
1. Draw your assumptions
1. Organize your data (what is time zero?)

---
class: title title-1

# Types of data

.box-1.medium[Cross sectional data]

--

.box-1.medium[Longitudional data]

---

class: title title-1

# Types of data

.box-1.medium[Cross sectional data]



.box-2.medium[Longitudional data]

---

class: title title-1

# Types of data

.box-1.medium[Cross sectional data]

.box-inv-1[One observation per person]

.box-2.medium[Longitudional data]

---

class: title title-1

# Types of data

.box-2.medium[Cross sectional data]

.box-2[One observation per person]

.box-1.medium[Longitudional data]

---
class: title title-1

# Causal inference

1. Ask an appropriate causal question
1. Draw your assumptions
1. Organize your data (what is time zero?)
1. Answer your causal question

---

class: title title-1

# Causal inference

1. Ask an appropriate causal question
1. Draw your assumptions
1. *Organize your data (what is time zero?)*
1. Answer your causal question

---
