---
title: "Law of Iterated Expectations"
author: "Lucy D'Agostino McGowan"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    css: ["default", "css/style.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```


```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view"))
```

class: center middle main-title section-title-1

# Law of Iterated Expectations

.class-info[

**Session 9**

.light[STA 379/679: Causal Inference <br>
Lucy D'Agostino McGowan
]

]

---

class: title title-1


# What is expectation?

.box-1.medium[
.huge[
$$E[X]$$
]
]

---

class: title title-1

# What is expectation

.box-1[In the discrete case, an expected value is just a weighted average of all possible values of the discrete random variable]

--
<br>
.box-1[

$$E[X] = \sum_xxP(X=x)$$

]
---

class: title title-1

# What is expectation

.box-1[How do we define expectation for continuous random variables?]

--

.box-1[

$$E[X] = \int_{-\infty}^\infty xf(x)dx$$

]

--
<br>
.box-inv-1[
What is $f(x)$?]

---


class: title title-1

# What is a conditional expectation?


.box-1[

$$E[X|Y] = \sum_xxP(X=x|Y=y)$$

]

---

class: title title-1

# What is a conditional expectation?


.box-1[

$$E[X|Y] = \sum_{\color{orange}x}xP(X=x|Y=y)$$

]

---

class: title title-1

# What is a conditional expectation?


.box-1[

$$E[X|Y] = \sum_{\color{orange}x}\color{orange}xP(X=x|Y=y)$$

]

---


class: title title-1

# What is a conditional expectation?


.box-1[

$$E[X|Y] = \sum_{\color{orange}x}\color{orange}x\color{orange}{P(X=x|Y=y)}$$

]
---

class: title title-1

# What is a conditional expectation?


.box-1[

$$E[X|Y] = \sum_{x}xP(X=x|Y=y)$$
]

<br>

.box-1[
$$E[X|Y] = \int_{-\infty}^\infty x f(x|y)dx$$
]

---


class: title title-1

# What is a conditional expectation?


.box-1[

$$E[X|Y] = \sum_{x}xP(X=x|Y=y)$$
]

<br>

.box-1[
$$E[X|Y] = \int_{-\infty}^\infty x \color{orange}{f(x|y)}dx$$
]
---

class: title title-1

# Law of Iterated Expectations

.box-1.medium[

$$\Large E[X]=E[E[X|Y]]$$
]

---

class: title title-1

# Law of Iterated Expectations

.box-1.medium[

$$\Large E_\color{orange}X[X]=E_\color{orange}Y[E[X|Y]]$$
]

---

class: title title-1

# Law of Iterated Expectations Proof

.box-1.medium[

$$\Large E[X]=E[E[X|Y]]$$
]

---


class: title title-1

# Law of Iterated Expectations Proof

.box-1.medium[

$$E[X]=E[\color{orange}{E[X|Y]}]$$
]


---


class: title title-1

# Law of Iterated Expectations Proof

.box-1.medium[

$$E[X]=E\left[\color{orange}{\sum_x xP(X=x|Y=y)}\right]$$
]

--

.box-inv-1[Definition of expectation]

---

class: title title-1

# Law of Iterated Expectations Proof

.box-1.medium[

$$E[X]=\color{orange}E\left[\sum_x xP(X=x|Y=y)\right]$$
]

---

class: title title-1

# Law of Iterated Expectations Proof

.box-1[

$$E[X]=\color{orange}{\sum_y}\sum_x xP(X=x|Y=y)\color{orange}{P(Y=y)}$$
]

--
<br>
.box-inv-1[Definition of expectation]

---

class: title title-1

# Law of Iterated Expectations Proof

.box-1[

$$E[X]=\sum_y\sum_x x\color{orange}{P(X=x|Y=y)P(Y=y)}$$
]

--
<br>

.box-1[

Per Bayes! $$P(X|Y)P(Y) = P(X,Y)$$

]

---

class: title title-1

# Law of Iterated Expectations Proof

.box-1[

$$E[X]=\sum_y\sum_x x\color{orange}{P(X=x, Y=y)}$$
]

---

class: title title-1

# Law of Iterated Expectations Proof

.box-1[

$$E[X]=\color{orange}{\sum_x \sum_y}xP(X=x, Y=y)$$
]

---


class: title title-1

# Law of Iterated Expectations Proof

.box-1[

$$E[X]=\sum_x x \sum_yP(X=x, Y=y)$$
]


---

class: title title-1

# Law of Iterated Expectations Proof

.box-1[

$$E[X]=\sum_x x \color{orange}{\sum_yP(X=x, Y=y)}$$
]

--
<br>
.box-1[
Probability! $\sum_yP(X, Y)=P(X)$
]

---

class: title title-1

# Law of Iterated Expectations Proof

.box-1.medium[

$$E[X]=\sum_x x \color{orange}{P(X = x)}$$
]

--
<br>

.box-1[Definition of expectations]

---

class: title title-1

# Law of Iterated Expectations Proof

.box-1.medium[

$$E[X]=E[X]$$
]

---

class: title title-1

# Law of Iterated Expectations Proof

.box-1[Suppose $X$ and $Y$ are jointly continuous with a joint density function $f(x, y)$ and marginal densities $f_x(x), f_y(y)$]

<br>

.box-1.medium[

$$E[X]=\int_{-\infty}^\infty \int_{-\infty}^\infty xf(x,y)dxdy$$
]

---

class: title title-1

# Law of Iterated Expectations Proof


.box-1.medium[

$$E[X]=\int_{-\infty}^\infty \int_{-\infty}^\infty xf(x,y)dxdy$$
]

---

class: title title-1

# Law of Iterated Expectations Proof


.box-1[

$$\begin{align}E[X]&=\int_{-\infty}^\infty \int_{-\infty}^\infty xf(x,y)dxdy\\&=\int_{-\infty}^\infty \int_{-\infty}^\infty xf(x|y)f_y(y)dxdy\end{align}$$
]

--

.box-1[Why?]

---

class: title title-1

# Law of Iterated Expectations Proof


.box-1[

$$\begin{align}E[X]&=\int_{-\infty}^\infty \int_{-\infty}^\infty xf(x,y)dxdy\\&=\int_{-\infty}^\infty \int_{-\infty}^\infty xf(x|y)f_y(y)dxdy\\&=\int_{-\infty}^\infty \left\{\int_{-\infty}^\infty xf(x|y)dx\right\}f_y(y)dy\end{align}$$
]

--

.box-1[What is in the brackets?]

---
class: title title-1

# Law of Iterated Expectations Proof


.box-1.small[

$$\begin{align}E[X]&=\int_{-\infty}^\infty \int_{-\infty}^\infty xf(x,y)dxdy\\&=\int_{-\infty}^\infty \int_{-\infty}^\infty xf(x|y)f_y(y)dxdy\\&=\int_{-\infty}^\infty \left\{\int_{-\infty}^\infty xf(x|y)dx\right\}f_y(y)dy\\&=\int_{-\infty}^\infty E[X|Y]f_y(y)dy\end{align}$$
]
---
class: title title-1

# Law of Iterated Expectations Proof


.box-1.small[

$$\begin{align}E[X]&=\int_{-\infty}^\infty \int_{-\infty}^\infty xf(x,y)dxdy\\&=\int_{-\infty}^\infty \int_{-\infty}^\infty xf(x|y)f_y(y)dxdy\\&=\int_{-\infty}^\infty \left\{\int_{-\infty}^\infty xf(x|y)dx\right\}f_y(y)dy\\&=\int_{-\infty}^\infty E[X|Y]f_y(y)dy\\&=E[E[X|Y]]\end{align}$$
]


---

class: title title-1

# Law of Iterated Expectations

.box-1.medium[
Practical use: if you know something about the conditional distribution $X|Y$, and you know something about $Y$, you can learn about $X$
]

---

class: title title-1

# Law of Iterated Expectations

.box-1.medium[Practical Example]

.box-inv-1.medium[We want to know the average exam grade for the class. We know:]

--
1. The average conditional on whether the student studied
2. The proportion of students who studied

---

class: title title-1

# Law of Iterated Expectations

.box-1.medium[Practical Example]

.box-inv-1.medium[We want to know the average exam grade for the class]



$$E[\textrm{exam grade}] = E[E[\textrm{exam grade | study status}]]$$
---

class: title title-1

# Law of Iterated Expectations

.box-1.medium[Practical Example]

.box-inv-1.medium[We want to know the average exam grade for the class]



$$E[\textrm{exam grade}] = \\\sum_{\textrm{study status}}E[\textrm{exam grade | study status}]P(\textrm{study status})$$

---

class: title title-1

# Law of Iterated Expectations

.box-1.medium[Practical Example]

.box-inv-1.medium[We want to know the average exam grade for the class]


$$E[\textrm{exam grade}] = \\E[\textrm{exam grade | studied}]P(\textrm{studied})+\\E[\textrm{exam grade | didn't study}]P(\textrm{didn't study})$$


---

class: title title-1

# Law of Iterated Expectations

.box-1.medium[Practical Example]

.box-inv-1.medium[We want to know the average exam grade for the class]
<br>
.box-1[
$E[\textrm{exam grade}] = 95 \times 0.8 + 70\times 0.2 = 90$
]

---

class: title title-1

# Example

Let's flip a coin twice, heads is 1, tails is 0. Let $X$ be the sum of the two flips, and $Y$ be the maximum. We can draw a table of the joint and marginal distributions:

$p_{X,Y}(x,y)$ | | |
----------------|--|-|
 | y = 0 | y = 1 | $p_X(x)$
x = 0 | 0.25 | 0 | 0.25
x = 1 | 0 | 0.5 | 0.5
x = 2 | 0 | 0.25 | 0.25
$p_Y(y)$|0.25 | 0.75


---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise



1. Using the table on the previous slide, calculate $E[Y]$
2. Calculate $E[Y|X = 0]$
3. Calculate $E[Y|X = x]$ for each possible value $x$ of $X$
4. Calculate $E[E[Y|X=x]]$ 

<br><br>
## https://bit.ly/sta-679-s22-ae7


---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

$p_{X,Y}(x,y)$ | | |
----------------|--|-|
 | y = 0 | y = 1 | $p_X(x)$
x = 0 | 0.25 | 0 | 0.25
x = 1 | 0 | 0.5 | 0.5
x = 2 | 0 | 0.25 | 0.25
$p_Y(y)$|0.25 | 0.75


.box-1[
$$E[Y] = \sum_y y P(Y = y)$$
]

---
class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

$p_{X,Y}(x,y)$ | | |
----------------|--|-|
 | y = 0 | y = 1 | $p_X(x)$
x = 0 | 0.25 | 0 | 0.25
x = 1 | 0 | 0.5 | 0.5
x = 2 | 0 | 0.25 | 0.25
$p_Y(y)$|0.25 | 0.75


.box-1[
$$E[Y] = 0 \times 0.25 + 1 \times 0.75 = 0.75$$


]

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

$p_{X,Y}(x,y)$ | | |
----------------|--|-|
 | y = 0 | y = 1 | $p_X(x)$
x = 0 | 0.25 | 0 | 0.25
x = 1 | 0 | 0.5 | 0.5
x = 2 | 0 | 0.25 | 0.25
$p_Y(y)$|0.25 | 0.75

.box-1[
$$E[Y|X = 0] = 0 \times 1 + 0 \times 0 = 0$$
]
---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.pull-left[
$p_{X,Y}(x,y)$ | | |
----------------|--|-|
 | y = 0 | y = 1 | $p_X(x)$
x = 0 | 0.25 | 0 | 0.25
x = 1 | 0 | 0.5 | 0.5
x = 2 | 0 | 0.25 | 0.25
$p_Y(y)$|0.25 | 0.75
]

.pull-right[
x | E[Y &#124; X = x]
---|--------
0 | $0\times 1 + 1\times 0 = 0$
1 | $0 \times 0 + 1 \times 1 = 1$
2 | $0 \times 0 + 1 \times 1 = 1$

]

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.pull-left[
$p_{X,Y}(x,y)$ | | |
----------------|--|-|
 | y = 0 | y = 1 | $p_X(x)$
x = 0 | 0.25 | 0 | 0.25
x = 1 | 0 | 0.5 | 0.5
x = 2 | 0 | 0.25 | 0.25
$p_Y(y)$|0.25 | 0.75
]

.pull-right[
x | E[Y &#124; X = x]
---|--------
0 | $0\times 1 + 1\times 0 = 0$
1 | $0 \times 0 + 1 \times 1 = 1$
2 | $0 \times 0 + 1 \times 1 = 1$

]

.box-1[
$$E[E[Y|X=x]] = 0 \times 0.25 + 1 \times 0.5 + 1 \times 0.25 = 0.75$$
]
---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise


```{r, eval = FALSE}
flip2 <- rbinom(100000, 1, 0.5)
flip1 <- rbinom(100000, 1, 0.5)
x <- flip2 + flip1
y = pmax(flip2, flip1)
mean(y)
mean(x)
mean(y[x == 0])
mean(y[x == 1])
mean(y[x == 2])
mean(y[x == 0]) * mean(x == 0) + 
  mean(y[x == 1]) * mean(x == 1) + 
  mean(y[x == 2]) * mean(x == 2)
```


