---
title: "Causal Estimands"
author: "Lucy D'Agostino McGowan"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    css: ["default", "css/style.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```


```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view"))
```

class: center middle main-title section-title-1

# Causal estimands

.class-info[

**Session 11**

.light[STA 379/679: Causal Inference <br>
Lucy D'Agostino McGowan
]

]

---

class: title title-1
# Average treatment effect

.box-1[
.huge[
$$ATE = E[Y(1) - Y(0)]$$
]
]

--

<br>

.box-inv-1[
Target population: everyone
]

---
class: title title-1

.small[
# Average treatment effect among the treated
]

.box-1[
$$ATT = E[Y(1) - Y(0) | X = 1]$$

]

--
<br>
.box-inv-1[
Target population: treated population
]

---
class: title title-1
.small[
# Average treatment effect among the controls
]

.box-1[
$$ATC = E[Y(1) - Y(0) | X = 0]$$
]


--
<br>
.box-inv-1[
Target population: control population
]

---
class: title title-1

.small[
# Average treatment effect among matchable
]

.box-1[
$$ATM_d = E[Y(1) - Y(0) | M_d = 1]$$
]

--
<br>

.box-inv-1[
Target population: matchable population
]

---
class: title title-1

.small[
# Average treatment effect among matchable
]


.box-1[
A subject is evenly matchable $(M_d = 1)$ for a given matching process, $d$
]

--

<br>

.box-1[
A subject is evenly matchable if the limit of the ratio of the number of subjects from the opposite treatment to the number from its own treatment is greater than 1 within the localized region of the covariate space around the subject defined by $d$
]

---

class: title title-1

.small[
# Average treatment effect among the overlap
]

.box-inv-1[
Target population: overlap population
]

--

.box-1[Similar to the "evenly matchable"]

--

.box-1[Creates a "pseudo-population" that has excellent variance properties]




