---
title: "Propensity score weighting"
author: "Lucy D'Agostino McGowan"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    css: ["default", "css/style.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```


```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view"))
```

class: center middle main-title section-title-1

# Propensity score weighting

.class-info[

**Session 11**

.light[STA 379/679: Causal Inference <br>
Lucy D'Agostino McGowan
]

]

---

class: title title-1
# Average treatment effect

.box-1[
$$ATE = E[Y(1) - Y(0)]$$
]

--

<br>


.box-inv-1[
$$w_{ATE} = \frac{X_i}{e_i}+\frac{1-X_i}{1-e_i}$$
]

---

class: title title-1
# Average treatment effect

```{r, include = FALSE}
library(broom)
library(tidyverse)
library(causaldata)
nhefs_complete_uc <- nhefs_complete %>% 
  filter(censored == 0)
```

```{r}
nhefs_complete_uc <-
  glm(qsmk ~ wt71 + age + marital + race + sex, 
      data = nhefs_complete_uc,
      family = binomial()) %>%
  augment(type.predict = "response",
          data = nhefs_complete_uc) 
```

---
class: title title-1
# Average treatment effect

```{r}
nhefs_complete_uc <- nhefs_complete_uc %>%
  mutate( 
    w_ate = qsmk / .fitted + 
      (1 - qsmk) / (1 - .fitted)
  )
```
---

class: title title-1
# Average treatment effect

```{r}
nhefs_complete_uc <- nhefs_complete_uc %>%
  mutate( #<<
    w_ate = qsmk / .fitted + 
      (1 - qsmk) / (1 - .fitted)
  )
```
---
class: title title-1
# Average treatment effect

```{r}
nhefs_complete_uc <- nhefs_complete_uc %>%
  mutate( 
    w_ate = qsmk / .fitted + #<<
      (1 - qsmk) / (1 - .fitted)
  )
```

---

class: title title-1
# Average treatment effect

```{r}
nhefs_complete_uc <- nhefs_complete_uc %>%
  mutate( 
    w_ate = qsmk / .fitted + 
      (1 - qsmk) / (1 - .fitted)#<<
  )
```

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.box-1[Open `appex-09` from last week]

.box-1[Calculate the ATE weights using the propensity score model you fit]

.box-1[What is the range of these weights? Theoretically, how small can they be? How large?]

.box-1[Knit, commit, push to GitHub]

`r countdown::countdown(5)`
---

class: title title-1

# Mirrored Histogram

.small[
```{r}
df_plot_ps <- nhefs_complete_uc %>%
  tidyr::pivot_wider(names_from = qsmk, 
                     values_from = .fitted, 
                     names_prefix = "qsmk_p")
```
]

---
class: title title-1
# Mirrored Histogram

.left-code[
.small[
```{r mirror2, eval = FALSE}
ggplot(df_plot_ps) +
  geom_histogram(bins = 50, 
                 aes(x = qsmk_p1), 
                 fill = "cornflower blue") + 
  geom_histogram(bins = 50, 
                 alpha = 0.5,
                 aes(x = qsmk_p1, 
                     weight = w_ate), 
                 fill = "cornflower blue") +
  geom_histogram(bins = 50, 
                 aes(x = qsmk_p0, 
                     y = - stat(count))) + 
  geom_histogram(bins = 50, 
                 alpha = 0.5,
                 aes(x = qsmk_p0, 
                     weight = w_ate,
                     y = - stat(count)), 
                 fill = "orange") + 
  scale_y_continuous("count", label = abs) + 
  scale_x_continuous("propensity score") +
  geom_label(
    label = "Quit smoking", 
    x = 0.25,
    y = 3,
  ) + 
  geom_label(
    label = "Didn't quit smoking",
    x = 0.25, 
    y = -3
  ) + 
  theme_minimal()
```
]
]

.right-plot[
```{r, ref.label="mirror2", echo=FALSE}
```
]
---
class: title title-1
# Mirrored Histogram

.left-code[
.small[
```{r, eval = FALSE}
ggplot(df_plot_ps) +
  geom_histogram(bins = 50, 
                 aes(x = qsmk_p1), 
                 fill = "cornflower blue") + 
  geom_histogram(bins = 50,  #<<
                 alpha = 0.5, #<<
                 aes(x = qsmk_p1, #<< 
                     weight = w_ate),  #<<
                 fill = "cornflower blue") +
  geom_histogram(bins = 50, 
                 aes(x = qsmk_p0, 
                     y = - stat(count))) + 
  geom_histogram(bins = 50, 
                 alpha = 0.5,
                 aes(x = qsmk_p0, 
                     weight = w_ate,
                     y = - stat(count)), 
                 fill = "orange") + 
  scale_y_continuous("count", label = abs) + 
  scale_x_continuous("propensity score") +
  geom_label(
    label = "Quit smoking", 
    x = 0.25,
    y = 3,
  ) + 
  geom_label(
    label = "Didn't quit smoking",
    x = 0.25, 
    y = -3
  ) + 
  theme_minimal()
```
]
]

.right-plot[
```{r, ref.label="mirror2", echo=FALSE}
```
]

---

class: title title-1
# Mirrored Histogram

.left-code[
.small[
```{r, eval = FALSE}
ggplot(df_plot_ps) +
  geom_histogram(bins = 50, 
                 aes(x = qsmk_p1), 
                 fill = "cornflower blue") + 
  geom_histogram(bins = 50,  
                 alpha = 0.5, 
                 aes(x = qsmk_p1, 
                     weight = w_ate),  
                 fill = "cornflower blue") +
  geom_histogram(bins = 50, 
                 aes(x = qsmk_p0, 
                     y = - stat(count))) + 
  geom_histogram(bins = 50, #<<
                 alpha = 0.5, #<<
                 aes(x = qsmk_p0, #<<
                     weight = w_ate,#<<
                     y = - stat(count)), #<<
                 fill = "orange") + 
  scale_y_continuous("count", label = abs) + 
  scale_x_continuous("propensity score") +
  geom_label(
    label = "Quit smoking", 
    x = 0.25,
    y = 3,
  ) + 
  geom_label(
    label = "Didn't quit smoking",
    x = 0.25, 
    y = -3
  ) + 
  theme_minimal()
```
]
]

.right-plot[
```{r, ref.label="mirror2", echo=FALSE}
```
]
---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.box-1[Create a mirrored histogram with your ATE pseudo-population overlaid]

.box-1[Knit, commit, push to GitHub]

`r countdown::countdown(5)`

---

class: title title-1

# Table 1!

.left-code[
.small[
```{r, eval = FALSE}
library(gtsummary)
nhefs_complete_uc %>%
  tbl_summary(
    by = qsmk, 
    include = c(
      "wt71", 
      "age", 
      "marital",
      "race",
      "sex")) %>%
  add_overall()
```

]]

.right-plot[
.small[
```{r, echo = FALSE}
library(gtsummary)
nhefs_complete_uc %>%
  tbl_summary(
    by = qsmk, 
    include = c("wt71", "age", "marital", "race", "sex")) %>%
  add_overall()
```
]
]
---

class: title title-1

# Table 1!

```{r, echo = FALSE}
nhefs_complete_uc %>%
  tbl_summary(
    by = qsmk, 
    include = c("wt71", "age", "marital", "race", "sex")) %>%
  add_overall()
```

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.box-1[`install.packages("gtsummary")`]

.box-1[Create Table 1 split by `qsmk`, including an `overall` column and using all variables in your propensity score model]

.box-1[Knit, commit, push to GitHub]

`r countdown::countdown(5)`

---

class: title title-1

# Weighted Table 1

```{r, eval = FALSE}
library(survey)
svy_des <- svydesign(
  ids = ~ 1,
  data = nhefs_complete_uc,
  weights = ~ w_ate
)
```

---

class: title title-1

# Weighted Table 1

```{r, eval = FALSE}
library(survey) #<<
svy_des <- svydesign(
  ids = ~ 1,
  data = nhefs_complete_uc,
  weights = ~ w_ate
)
```

---

class: title title-1

# Weighted Table 1

```{r, eval = FALSE}
library(survey)
svy_des <- svydesign( #<<
  ids = ~ 1,
  data = nhefs_complete_uc,
  weights = ~ w_ate
)
```

---

class: title title-1

# Weighted Table 1

```{r, eval = FALSE}
library(survey)
svy_des <- svydesign(
  ids = ~ 1, #<<
  data = nhefs_complete_uc,
  weights = ~ w_ate
)
```

---

class: title title-1

# Weighted Table 1

```{r, eval = FALSE}
library(survey)
svy_des <- svydesign(
  ids = ~ 1,
  data = nhefs_complete_uc, #<<
  weights = ~ w_ate
)
```

---

class: title title-1

# Weighted Table 1

```{r}
library(survey)
svy_des <- svydesign(
  ids = ~ 1,
  data = nhefs_complete_uc,
  weights = ~ w_ate #<<
)
```

---


class: title title-1

# Weighted Table 1

```{r, eval = FALSE}
svy_des %>%
  tbl_svysummary(by = qsmk,
              include = c("wt71",
                          "age",
                          "marital",
                          "race",
                          "sex")) %>%
  add_overall()
```

---

class: title title-1

# Weighted Table 1

```{r, eval = FALSE}
svy_des %>% #<<
  tbl_svysummary(by = qsmk, 
              include = c("wt71",
                          "age",
                          "marital",
                          "race",
                          "sex")) %>%
  add_overall()
```
---

class: title title-1

# Weighted Table 1

```{r, eval = FALSE}
svy_des %>%
  tbl_svysummary(by = qsmk, #<<
              include = c("wt71",
                          "age",
                          "marital",
                          "race",
                          "sex")) %>%
  add_overall()
```

---

class: title title-1

# Weighted Table 1

```{r, echo = FALSE}
svy_des %>%
  tbl_svysummary(by = qsmk,
              include = c("wt71",
                          "age",
                          "marital",
                          "race",
                          "sex")) %>%
  add_overall()
```

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.box-1[`install.packages("survey")`]

.box-1[Create weighted Table 1. Compare this to your Table 1 from the previous exercise - which column does it resemble the most?]

.box-1[Knit, commit, push to GitHub]

`r countdown::countdown(5)`

---

class: title title-1

.small[
# Average treatment effect among the treated
]

.box-1[
$$ATT = E[Y(1) - Y(0) | X = 1]$$

]

--

<br>

.box-1[

$$w_{ATT} = \frac{e_i X_i}{e_i} + \frac{e_i(1-X_i)}{1-e_i}$$
]

--
<br>
.box-inv-1[What is this when X=1?]
---
class: title title-1

# ATT

```{r, echo = FALSE}
df_plot_ps <- nhefs_complete_uc %>%
  mutate(w_att = qsmk + .fitted * (1 - qsmk) / (1 - .fitted) ) %>%
  tidyr::pivot_wider(names_from = qsmk, 
                     values_from = .fitted, 
                     names_prefix = "qsmk_p")
ggplot(df_plot_ps) +
  geom_histogram(bins = 50, 
                 aes(x = qsmk_p1), 
                 fill = "cornflower blue") + 
  geom_histogram(bins = 50,  
                 alpha = 0.5, 
                 aes(x = qsmk_p1, 
                     weight = w_att),  
                 fill = "cornflower blue") +
  geom_histogram(bins = 50, 
                 aes(x = qsmk_p0, 
                     y = - stat(count))) + 
  geom_histogram(bins = 50, #<<
                 alpha = 0.5, #<<
                 aes(x = qsmk_p0, #<<
                     weight = w_att,#<<
                     y = - stat(count)), #<<
                 fill = "orange") + 
  scale_y_continuous("count", label = abs) + 
  scale_x_continuous("propensity score") +
  geom_label(
    label = "Quit smoking", 
    x = 0.25,
    y = 3,
  ) + 
  geom_label(
    label = "Didn't quit smoking",
    x = 0.25, 
    y = -3
  ) + 
  theme_minimal()
```

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.box-1[Calculate the `w_att`. What is the range of possible weights?]

.box-1[Plot a mirrored histogram of this pseudo-population. What do you notice?]

.box-1[Create weighted Table 1. Compare this to your unweighted Table 1 - which column does it resemble the most?]

.box-1[Knit, commit, push to GitHub]

`r countdown::countdown(10)`

---
class: title title-1
.small[
# Average treatment effect among the controls
]

.box-1[
$$ATC = E[Y(1) - Y(0) | X = 0]$$
]


--
<br>

.box-1[

$$w_{ATC} = \frac{(1 - e_i) X_i}{e_i} + \frac{(1 - e_i)(1-X_i)}{1-e_i}$$
]

--

<br>
.box-inv-1[What is this when X=0?]

---

class: title title-1

# ATC

```{r, echo = FALSE}
df_plot_ps <- nhefs_complete_uc %>%
  mutate(w_atc = (1 - .fitted) * qsmk / .fitted + (1 - qsmk)) %>%
  tidyr::pivot_wider(names_from = qsmk, 
                     values_from = .fitted, 
                     names_prefix = "qsmk_p")
ggplot(df_plot_ps) +
  geom_histogram(bins = 50, 
                 aes(x = qsmk_p1), 
                 fill = "cornflower blue") + 
  geom_histogram(bins = 50,  
                 alpha = 0.5, 
                 aes(x = qsmk_p1, 
                     weight = w_atc),  
                 fill = "cornflower blue") +
  geom_histogram(bins = 50, 
                 aes(x = qsmk_p0, 
                     y = - stat(count))) + 
  geom_histogram(bins = 50, #<<
                 alpha = 0.5, #<<
                 aes(x = qsmk_p0, #<<
                     weight = w_atc,#<<
                     y = - stat(count)), #<<
                 fill = "orange") + 
  scale_y_continuous("count", label = abs) + 
  scale_x_continuous("propensity score") +
  geom_label(
    label = "Quit smoking", 
    x = 0.25,
    y = 3,
  ) + 
  geom_label(
    label = "Didn't quit smoking",
    x = 0.25, 
    y = -3
  ) + 
  theme_minimal()
```

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.box-1[Calculate the `w_atc`. What is the range of possible weights?]

.box-1[Plot a mirrored histogram of this pseudo-population. What do you notice?]

.box-1[Create weighted Table 1. Compare this to your unweighted Table 1 - which column does it resemble the most?]

.box-1[Knit, commit, push to GitHub]

`r countdown::countdown(10)`

---
class: title title-1

.small[
# Average treatment effect among matchable
]

.box-1[
$$ATM_d = E[Y(1) - Y(0) | M_d = 1]$$
]

--
<br>

.box-inv-1[

$$w_{ATM} = \frac{\textrm{min}\{e_i, 1-e_i\}}{X_i e_i + (1 - X_i)(1 - e_i)}$$
]

---

class: title title-1
.small[
# Average treatment effect among matchable
]

```{r}
nhefs_complete_uc <- nhefs_complete_uc %>%
  mutate( 
    w_atm = pmin(.fitted, 1 - .fitted) / #<<
      (qsmk * .fitted + (1 - qsmk) * (1 - .fitted)) #<<
  )
```

---

class: title title-1

# ATM

```{r, echo = FALSE}
df_plot_ps <- nhefs_complete_uc %>%
  tidyr::pivot_wider(names_from = qsmk, 
                     values_from = .fitted, 
                     names_prefix = "qsmk_p")
ggplot(df_plot_ps) +
  geom_histogram(bins = 50, 
                 aes(x = qsmk_p1), 
                 fill = "cornflower blue") + 
  geom_histogram(bins = 50,  
                 alpha = 0.5, 
                 aes(x = qsmk_p1, 
                     weight = w_atm),  
                 fill = "cornflower blue") +
  geom_histogram(bins = 50, 
                 aes(x = qsmk_p0, 
                     y = - stat(count))) + 
  geom_histogram(bins = 50, #<<
                 alpha = 0.5, #<<
                 aes(x = qsmk_p0, #<<
                     weight = w_atm,#<<
                     y = - stat(count)), #<<
                 fill = "orange") + 
  scale_y_continuous("count", label = abs) + 
  scale_x_continuous("propensity score") +
  geom_label(
    label = "Quit smoking", 
    x = 0.25,
    y = 3,
  ) + 
  geom_label(
    label = "Didn't quit smoking",
    x = 0.25, 
    y = -3
  ) + 
  theme_minimal()
```

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.box-1[Calculate the `w_atm`. What is the range of possible weights?]

.box-1[Plot a mirrored histogram of this pseudo-population. What do you notice?]

.box-1[Create weighted Table 1. Compare this to your unweighted Table 1 - which column does it resemble the most (if any)?]

.box-1[Knit, commit, push to GitHub]

`r countdown::countdown(10)`

---
class: title title-1

.small[
# Average treatment effect among the overlap
]

.box-inv-1[
Target population: overlap population
]

--

.box-1[Similar to the "evenly matchable"]

--

.box-1[Creates a "pseudo-population" that has excellent variance properties]

--

.box-1[
$$w_{ATO} = (1-e_i)X_i + e_i (1-X_i)$$
]
---

class: title title-1

.small[
# Average treatment effect among the overlap
]


.box-1[
$$w_{ATO} = (1-e_i)X_i + e_i (1-X_i)$$
]

--
<br>
.box-inv-1[
When the conditional probability of receiving the exposure is small, what does this approximate?
]

--
<br>
.box-1[

$$e_i\approx 0$$

$w_{ATT}$
]

---

class: title title-1

.small[
# Average treatment effect among the overlap
]


.box-1[
$$w_{ATO} = (1-e_i)X_i + e_i (1-X_i)$$
]

--
<br>
.box-1[

$$e_i\approx 1$$

$w_{ATC}$
]

---
class: title title-1

# ATO

```{r, echo = FALSE}
df_plot_ps <- nhefs_complete_uc %>%
  mutate(w_ato = (1 - .fitted) * qsmk + .fitted * (1 - qsmk)) %>%
  tidyr::pivot_wider(names_from = qsmk, 
                     values_from = .fitted, 
                     names_prefix = "qsmk_p")
ggplot(df_plot_ps) +
  geom_histogram(bins = 50, 
                 aes(x = qsmk_p1), 
                 fill = "cornflower blue") + 
  geom_histogram(bins = 50,  
                 alpha = 0.5, 
                 aes(x = qsmk_p1, 
                     weight = w_ato),  
                 fill = "cornflower blue") +
  geom_histogram(bins = 50, 
                 aes(x = qsmk_p0, 
                     y = - stat(count))) + 
  geom_histogram(bins = 50, #<<
                 alpha = 0.5, #<<
                 aes(x = qsmk_p0, #<<
                     weight = w_ato,#<<
                     y = - stat(count)), #<<
                 fill = "orange") + 
  scale_y_continuous("count", label = abs) + 
  scale_x_continuous("propensity score") +
  geom_label(
    label = "Quit smoking", 
    x = 0.25,
    y = 3,
  ) + 
  geom_label(
    label = "Didn't quit smoking",
    x = 0.25, 
    y = -3
  ) + 
  theme_minimal()
```

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.box-1[Calculate the `w_ato`. What is the range of possible weights?]

.box-1[Plot a mirrored histogram of this pseudo-population. What do you notice?]

.box-1[Create weighted Table 1. Compare this to your unweighted Table 1 - which column does it resemble the most (if any)?]

.box-1[Knit, commit, push to GitHub]

`r countdown::countdown(10)`