---
title: "Propensity score matching"
author: "Lucy D'Agostino McGowan"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    css: ["default", "css/style.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```


```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view"))
```

class: center middle main-title section-title-1

# Propensity score matching

.class-info[

**Session 12**

.light[STA 379/679: Causal Inference <br>
Lucy D'Agostino McGowan
]

]

---

class: title title-1

# Matching: ATT

```{r, include = FALSE}
library(broom)
library(tidyverse)
library(causaldata)
nhefs_complete_uc <- nhefs_complete %>% 
  filter(censored == 0)
```

.small[
```{r, eval = FALSE}
library(MatchIt)
m <- matchit(qsmk ~ wt71 + age + marital + race + sex, 
  data = nhefs_complete_uc)
m
```
]

---
class: title title-1

# Matching: ATT

.small[
```{r, eval = FALSE}
library(MatchIt) #<<
m <- matchit(qsmk ~ wt71 + age + marital + race + sex, 
  data = nhefs_complete_uc)
m
```
]

---


class: title title-1

# Matching: ATT

.small[
```{r, eval = FALSE}
library(MatchIt)
m <- matchit(qsmk ~ wt71 + age + marital + race + sex, #<<
  data = nhefs_complete_uc)
m
```
]

---
class: title title-1

# Matching: ATT

.small[
```{r}
library(MatchIt)
m <- matchit(qsmk ~ wt71 + age + marital + race + sex, 
  data = nhefs_complete_uc)
m
```
]
---
class: title title-1

# Matching: ATT

.small[
```{r, highlight.output = 6}
library(MatchIt)
m <- matchit(qsmk ~ wt71 + age + marital + race + sex, 
  data = nhefs_complete_uc)
m
```
]
---

class: title title-1

# Matching: ATT

.small[
```{r, eval = FALSE}
matched_data <- get_matches(m, id = "i")
glimpse(matched_data)
```
]
---
class: title title-1

# Matching: ATT

.small[
```{r, eval = FALSE}
matched_data <- get_matches(m, id = "i") #<<
glimpse(matched_data)
```
]

---

class: title title-1

# Matching: ATT

.small[
```{r, eval = FALSE}
matched_data <- get_matches(m, id = "i")
glimpse(matched_data) #<<
```
]

---
class: title title-1

# Matching: ATT

.small[
```{r}
matched_data <- get_matches(m, id = "i")
glimpse(matched_data)
```
]

---

class: title title-1

# Matching: ATT

.small[
```{r, highlight.output = 1}
matched_data <- get_matches(m, id = "i")
glimpse(matched_data)
```
]

---

class: title title-1

# Matching: ATT

.small[
```{r, highlight.output = 2}
matched_data <- get_matches(m, id = "i")
glimpse(matched_data)
```
]

---
class: title title-1

# Matching: ATT

.small[
```{r, highlight.output = 3}
matched_data <- get_matches(m, id = "i")
glimpse(matched_data)
```
]
---
class: title title-1

# Matching: ATT

.small[
```{r, highlight.output = 4}
matched_data <- get_matches(m, id = "i")
glimpse(matched_data)
```
]
---


class: title title-1

# Matching: ATC

.small[
```{r, eval = FALSE}
library(MatchIt)
m <- matchit(qsmk ~ wt71 + age + marital + race + sex, 
  data = nhefs_complete_uc,
  estimand = "ATC")
m
```
]

---

class: title title-1

# Matching: ATC

.small[
```{r, eval = FALSE}
library(MatchIt)
m <- matchit(qsmk ~ wt71 + age + marital + race + sex, 
  data = nhefs_complete_uc,
  estimand = "ATC") #<<
m
```
]

---

class: title title-1

# Matching: ATC

.small[
```{r}
library(MatchIt)
m <- matchit(qsmk ~ wt71 + age + marital + race + sex, 
  data = nhefs_complete_uc,
  estimand = "ATC")
m
```
]
---

class: title title-1

# Matching: ATC

.small[
```{r, highlight.output = 6}
library(MatchIt)
m <- matchit(qsmk ~ wt71 + age + marital + race + sex, 
  data = nhefs_complete_uc,
  estimand = "ATC")
m
```
]

---

class: title title-1

# Matching ATM

.small[
```{r, eval = FALSE}
library(MatchIt)
m <- matchit(qsmk ~ wt71 + age + marital + race + sex, 
  data = nhefs_complete_uc,
  link = "linear.logit",
  caliper = 0.1)
m
```
]

---

class: title title-1

# Matching ATM

.small[
```{r, eval = FALSE}
library(MatchIt)
m <- matchit(qsmk ~ wt71 + age + marital + race + sex, 
  data = nhefs_complete_uc,
  link = "linear.logit", #<<
  caliper = 0.1)
m
```
]

---

class: title title-1

# Matching ATM

.small[
```{r, eval = FALSE}
library(MatchIt)
m <- matchit(qsmk ~ wt71 + age + marital + race + sex, 
  data = nhefs_complete_uc,
  link = "linear.logit",
  caliper = 0.1) #<<
m
```
]

--

.box-1[Only matches propensity scores withing 0.1 standard deviations on the linear-logit scale]

---
class: title title-1

# Matching ATM

.small[
```{r}
library(MatchIt)
m <- matchit(qsmk ~ wt71 + age + marital + race + sex, 
  data = nhefs_complete_uc,
  link = "linear.logit",
  caliper = 0.1) 
m
```
]

---

class: title title-1
# Matching ATM


.small[
```{r, highlight.output = 7}
library(MatchIt)
m <- matchit(qsmk ~ wt71 + age + marital + race + sex, 
  data = nhefs_complete_uc,
  link = "linear.logit",
  caliper = 0.1) 
m
```
]

---

class: title title-1

# Matching ATM

.small[
```{r}
matched_data <- get_matches(m, id = "i")
glimpse(matched_data)
```
]

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.box-1[Open `appex-09` from last class]

.box-1[Using the same propensity score model, calculate ATM matches with a caliper of 0.05]

.box-1[Extract your matched data set and create a Table 1 of just this data]

.box-1[Knit, commit, push to GitHub]

`r countdown::countdown(10)`

