---
title: "Evaluating your propensity score model"
author: "Lucy D'Agostino McGowan"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    css: ["default", "css/style.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```


```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view"))
```

class: center middle main-title section-title-1

# Evaluating your propensity score model

.class-info[

**Session 14**

.light[STA 379/679: Causal Inference <br>
Lucy D'Agostino McGowan
]

]

---
class: title title-1

# Checking balance

.box-1.medium[Love Plots]

--

.box-1.medium[eCDF plots]

---
class: title title-1

# Standardized Mean Difference (SMD)

$$\LARGE d = \frac{\bar{z}_{exposed}-\bar{z}_{unexposed}}{\sqrt{\frac{s^2_{exposed}+s^2_{unexposed}}{2}}}$$
--
<br>
.box-1[Rule of thumb: SMD < 0.1]
---

class: title title-1

# Customer satisfaction data

```{r, include = FALSE}
library(tidyverse)
customer_satisfaction <- read_csv("../lab-02-propensity-score/customer_satisfaction.csv")
```

.small[
```{r, message = FALSE, warning = FALSE, eval = FALSE}
library(tidyverse)
library(broom)

glm(satisfied_customer_service ~ income + married + n_kids + has_pets + age + 
      education + gender + former_customer_service + previous_spend, 
    family = binomial(),
    data = customer_satisfaction) %>%
  augment(data = customer_satisfaction, type.predict = "response") %>%
  mutate(atm_wts = pmin(.fitted, 1 - .fitted) / 
           (satisfied_customer_service * .fitted + 
              (1 - satisfied_customer_service) * (1 - .fitted))
  ) -> customer_satisfaction
```
]

---

class: title title-1

# Customer satisfaction data

.small[
```{r, message = FALSE, warning = FALSE, eval = FALSE}
library(tidyverse)
library(broom)

glm(satisfied_customer_service ~ income + married + n_kids + has_pets + age + 
      education + gender + former_customer_service + previous_spend, 
        family = binomial(),
    data = customer_satisfaction) %>%
  augment(data = customer_satisfaction, type.predict = "response") %>%
  mutate(atm_wts = pmin(.fitted, 1 - .fitted) / 
           (satisfied_customer_service * .fitted + 
              (1 - satisfied_customer_service) * (1 - .fitted))
  ) -> customer_satisfaction
```
]

.box-1[Which confounders did I identify?]

---

class: title title-1

# Customer satisfaction data

.small[
```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(broom)

glm(satisfied_customer_service ~ income + married + n_kids + has_pets + age + #<<
      education + gender + former_customer_service + previous_spend, #<<
    family = binomial(),
    data = customer_satisfaction) %>%
  augment(data = customer_satisfaction, type.predict = "response") %>%
  mutate(atm_wts = pmin(.fitted, 1 - .fitted) / 
           (satisfied_customer_service * .fitted + 
              (1 - satisfied_customer_service) * (1 - .fitted))
  ) -> customer_satisfaction
```
]
---

class: title title-1

# Customer satisfaction data

.small[
```{r, message = FALSE, warning = FALSE, eval = FALSE}
library(tidyverse)
library(broom)

glm(satisfied_customer_service ~ income + married + n_kids + has_pets + age + 
      education + gender + former_customer_service + previous_spend, 
    family = binomial(),
    data = customer_satisfaction) %>%
  augment(data = customer_satisfaction, type.predict = "response") %>%
  mutate(atm_wts = pmin(.fitted, 1 - .fitted) / 
           (satisfied_customer_service * .fitted + 
              (1 - satisfied_customer_service) * (1 - .fitted))
  ) -> customer_satisfaction
```
]

.box-1[What estimand am I interested in?]
---
class: title title-1
# Customer satisfaction data

.small[
```{r, message = FALSE, warning = FALSE, eval = FALSE}
library(tidyverse)
library(broom)

glm(satisfied_customer_service ~ income + married + n_kids + has_pets + age + 
      education + gender + former_customer_service + previous_spend, 
    family = binomial(),
    data = customer_satisfaction) %>%
  augment(data = customer_satisfaction, type.predict = "response") %>%
  mutate(atm_wts = pmin(.fitted, 1 - .fitted) /  #<<
           (satisfied_customer_service * .fitted + #<<
              (1 - satisfied_customer_service) * (1 - .fitted)) #<<
  ) -> customer_satisfaction
```
]

.box-1[What estimand am I interested in?]
---

class: title title-1

# SMD in R

```{r, message = FALSE, warning = FALSE, eval = FALSE}
library(smd)
library(tidyverse)
df %>% 
  # w is optional
  summarise(smd = smd(confounder_1, 
                      exposure, 
                      w = wts)$estimate)
```

---
class: title title-1

# SMD in R

.small[
```{r, message = FALSE, warning = FALSE}
library(smd)
library(tidyverse)
customer_satisfaction %>% 
  # w is optional
  summarize(smd = smd(income, 
                      satisfied_customer_service, 
                      w = atm_wts)$estimate)
```
]


---

class: title title-1

# SMD in R

```{r,  message = FALSE, warning = FALSE, eval = FALSE}
smds <- df %>% 
  summarise( 
    across( 
      c(confounder_1, confounder_2, ...), 
      list(
        unweighted = ~smd(.x, exposure)$estimate, 
        weighted = ~smd(.x, exposure, wts)$estimate 
      )
    )
  )
```
---

class: title title-1

# SMD in R

```{r,  message = FALSE, warning = FALSE, eval = FALSE}
smds <- df %>% 
  summarise( 
    across( 
      c(confounder_1, confounder_2, ...),  #<<
      list(
        unweighted = ~smd(.x, exposure)$estimate, 
        weighted = ~smd(.x, exposure, wts)$estimate 
      )
    )
  )
```

---

class: title title-1

# SMD in R

```{r,  message = FALSE, warning = FALSE, eval = FALSE}
smds <- df %>% 
  summarise( 
    across( 
      c(confounder_1, confounder_2, ...), 
      list(
        unweighted = ~smd(.x, exposure)$estimate, #<<
        weighted = ~smd(.x, exposure, wts)$estimate #<<
      )
    )
  )
```

---

class: title title-1

# SMD in R

.small[
```{r,  message = FALSE, warning = FALSE}
smds <- customer_satisfaction %>% 
  summarise( 
    across( 
      c(income, married, n_kids, has_pets, age, 
        education, gender, former_customer_service, previous_spend), 
      list(
        unweighted = ~smd(.x, satisfied_customer_service)$estimate, 
        weighted = ~smd(.x, satisfied_customer_service, atm_wts)$estimate 
      )
    )
  )
```
]

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.box-1[Go to our Github organization and open `appex-10`]

.box-1[Fit a propensity score model for `satisfied_customer_service` (you can use the one you chose for Lab 2)]

.box-1[Calculate the `ate` and `ato` weights]

.box-1[Calculate the standardized mean differences (unweighted, `ate` weighted and `ato` weighted) for all of the variables (excluding the outcome, `next_spend`)]

.box-1[Knit, commit, push to GitHub]

`r countdown::countdown(10)`
---

class: title title-1

# SMD in R

.small[
```{r}
smds
```
]

---

class: title title-1

# SMD in R

```{r, eval = FALSE}
plot_df <- smds %>% 
  pivot_longer( 
    everything(),
    values_to = "SMD", 
    names_to = c("variable", "Method"), 
    names_pattern = "(.*)_(.*)"
  )
```

---

class: title title-1

# SMD in R

```{r, eval = FALSE}
plot_df <- smds %>% 
  pivot_longer( #<<
    everything(),
    values_to = "SMD", 
    names_to = c("variable", "Method"), 
    names_pattern = "(.*)_(.*)"
  )
```

---

class: title title-1

# SMD in R

```{r, eval = FALSE}
plot_df <- smds %>% 
  pivot_longer( 
    everything(), #<<
    values_to = "SMD", 
    names_to = c("variable", "Method"), 
    names_pattern = "(.*)_(.*)"
  )
```

---

class: title title-1

# SMD in R

```{r, eval = FALSE}
plot_df <- smds %>% 
  pivot_longer( 
    everything(),
    values_to = "SMD",  #<<
    names_to = c("variable", "Method"), 
    names_pattern = "(.*)_(.*)"
  )
```

---
class: title title-1

# SMD in R

```{r}
plot_df <- smds %>% 
  pivot_longer( 
    everything(),
    values_to = "SMD", 
    names_to = c("variable", "Method"), #<<
    names_pattern = "(.*)_(.*)"#<<
  )
```

---
class: title title-1

# SMD in R

```{r}
plot_df <- smds %>% 
  pivot_longer( 
    everything(),
    values_to = "SMD", 
    names_to = c("variable", "Method"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  arrange(Method, abs(SMD)) %>%
  mutate(variable = fct_inorder(variable))
```


---

class: title title-1

# SMD in R

.small[
```{r}
plot_df
```
]

---

class: title title-1

# Love Plot

.left-code[
```{r, eval = FALSE}
ggplot(
  data = plot_df,
  aes(x = abs(SMD), y = variable, 
      group = Method, color = Method)
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1, 
             color = "black", size = 0.1)
```
]

---

class: title title-1

# Love Plot

.left-code[
```{r, eval = FALSE}
ggplot(
  data = plot_df,
  aes(x = abs(SMD), y = variable, #<<
      group = Method, color = Method) #<<
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1, 
             color = "black", size = 0.1)
```
]

---

class: title title-1

# Love Plot

.left-code[
```{r, eval = FALSE}
ggplot(
  data = plot_df,
  aes(x = abs(SMD), y = variable, 
      group = Method, color = Method)
) +  
  geom_line(orientation = "y") + #<<
  geom_point() + 
  geom_vline(xintercept = 0.1, 
             color = "black", size = 0.1)
```
]

---

class: title title-1

# Love Plot

.left-code[
```{r, eval = FALSE}
ggplot(
  data = plot_df,
  aes(x = abs(SMD), y = variable, 
      group = Method, color = Method)
) +  
  geom_line(orientation = "y") +
  geom_point() + #<<
  geom_vline(xintercept = 0.1, 
             color = "black", size = 0.1)
```
]

---

class: title title-1

# Love Plot

.left-code[
```{r, eval = FALSE}
ggplot(
  data = plot_df,
  aes(x = abs(SMD), y = variable, 
      group = Method, color = Method)
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1,  #<<
             color = "black", size = 0.1) #<<
```
]

---

class: title title-1

# Love Plot

.left-code[
```{r, eval = FALSE}
ggplot(
  data = plot_df,
  aes(x = abs(SMD), y = variable, 
      group = Method, color = Method)
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1,  
             color = "black", size = 0.1) 
```

]

.right-plot[

```{r, echo = FALSE}
ggplot(
  data = plot_df,
  aes(x = abs(SMD), y = variable, 
      group = Method, color = Method)
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1,  #<<
             color = "black", size = 0.1) #<<
```

]
---
class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.box-1[Pivot your standardized mean difference output to make it plot-able]

.box-1[Arrange your data frame by method and (absolute) standardized mean difference]

.box-1[Create a Love Plot of your standardized mean differences]

.box-1[Knit, commit, push to GitHub]

`r countdown::countdown(10)`

---


class: title title-1

# Compare Matching

.small[
```{r, eval = FALSE}
library(MatchIt)
matched_smds <- matchit(satisfied_customer_service ~ income + married +  
                          n_kids + has_pets + age + education + gender + 
                          former_customer_service + previous_spend, 
                        data = customer_satisfaction, caliper = 0.5) %>%
  get_matches() %>%
  summarise( 
    across( 
      c(income, married, n_kids, has_pets, age, 
        education, gender, former_customer_service, previous_spend), 
      list(
        matched = ~smd(.x, satisfied_customer_service)$estimate
      )
    )
  )
```
]
---

class: title title-1

# Compare Matching

.small[
```{r, eval = FALSE}
library(MatchIt)#<<
matched_smds <- matchit(satisfied_customer_service ~ income + married +  
                          n_kids + has_pets + age + education + gender + 
                          former_customer_service + previous_spend, 
                        data = customer_satisfaction, caliper = 0.5) %>%
  get_matches() %>%
  summarise( 
    across( 
      c(income, married, n_kids, has_pets, age, 
        education, gender, former_customer_service, previous_spend), 
      list(
        matched = ~smd(.x, satisfied_customer_service)$estimate
      )
    )
  )
```
]

---

class: title title-1

# Compare Matching

.small[
```{r, eval = FALSE}
library(MatchIt)
matched_smds <- matchit(satisfied_customer_service ~ income + married +  #<<
                          n_kids + has_pets + age + education + gender + #<<
                          former_customer_service + previous_spend, #<<
                        data = customer_satisfaction, caliper = 0.5) %>%
  get_matches() %>%
  summarise( 
    across( 
      c(income, married, n_kids, has_pets, age, 
        education, gender, former_customer_service, previous_spend), 
      list(
        matched = ~smd(.x, satisfied_customer_service)$estimate
      )
    )
  )
```
]

---

class: title title-1

# Compare Matching

.small[
```{r, eval = FALSE}
library(MatchIt)
matched_smds <- matchit(satisfied_customer_service ~ income + married +  
                          n_kids + has_pets + age + education + gender + 
                          former_customer_service + previous_spend, 
                        data = customer_satisfaction, caliper = 0.5) %>% #<<
  get_matches() %>%
  summarise( 
    across( 
      c(income, married, n_kids, has_pets, age, 
        education, gender, former_customer_service, previous_spend), 
      list(
        matched = ~smd(.x, satisfied_customer_service)$estimate
      )
    )
  )
```
]

---

class: title title-1

# Compare Matching

.small[
```{r, eval = FALSE}
library(MatchIt)
matched_smds <- matchit(satisfied_customer_service ~ income + married +  
                          n_kids + has_pets + age + education + gender + 
                          former_customer_service + previous_spend, 
                        data = customer_satisfaction, caliper = 0.5) %>%
  get_matches() %>% #<<
  summarise( 
    across( 
      c(income, married, n_kids, has_pets, age, 
        education, gender, former_customer_service, previous_spend), 
      list(
        matched = ~smd(.x, satisfied_customer_service)$estimate
      )
    )
  )
```
]

---

class: title title-1

# Compare Matching

.small[
```{r}
library(MatchIt)
matched_smds <- matchit(satisfied_customer_service ~ income + married +  
                          n_kids + has_pets + age + education + gender + 
                          former_customer_service + previous_spend, 
                        data = customer_satisfaction, caliper = 0.5) %>%
  get_matches() %>%
  summarise( #<<
    across( #<<
      c(income, married, n_kids, has_pets, age, #<<
        education, gender, former_customer_service, previous_spend), #<<
      list(#<<
        matched = ~smd(.x, satisfied_customer_service)$estimate#<<
      )#<<
    )#<<
  )#<<
```
]

---

class: title title-1

# Compare Matching

.small[
```{r}
matched_smds
```
]

---

class: title title-1

# Compare Matching

```{r, eval = FALSE}
plot_df_all <- matched_smds %>%
  pivot_longer( 
    everything(),
    values_to = "SMD", 
    names_to = c("variable", "Method"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  bind_rows(plot_df)
```

---

class: title title-1

# Compare Matching

```{r, eval = FALSE}
plot_df_all <- matched_smds %>% #<<
  pivot_longer( 
    everything(),
    values_to = "SMD", 
    names_to = c("variable", "Method"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  bind_rows(plot_df)
```

---

class: title title-1

# Compare Matching

```{r, eval = FALSE}
plot_df_all <- matched_smds %>% 
  pivot_longer(  #<<
    everything(), #<<
    values_to = "SMD", #<<
    names_to = c("variable", "Method"),#<<
    names_pattern = "(.*)_(.*)"#<<
  ) %>%
  bind_rows(plot_df)
```

---

class: title title-1

# Compare Matching

```{r, eval = FALSE}
plot_df_all <- matched_smds %>% 
  pivot_longer( 
    everything(),
    values_to = "SMD", 
    names_to = c("variable", "Method"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  bind_rows(plot_df) #<<
```

---

class: title title-1

# Compare Matching

```{r}
plot_df_all <- matched_smds %>% 
  pivot_longer( 
    everything(),
    values_to = "SMD", 
    names_to = c("variable", "Method"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  bind_rows(plot_df) %>%
  arrange(Method, abs(SMD)) %>%
  mutate(variable = fct_inorder(variable))
```

---

class: title title-1

# Love Plot

.left-code[
```{r, eval = FALSE}
ggplot(
  data = plot_df_all,
  aes(x = abs(SMD), y = variable, 
      group = Method, color = Method)
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1,  
             color = "black", size = 0.1) 
```
]

.right-plot[
```{r, echo = FALSE}
ggplot(
  data = plot_df_all,
  aes(x = abs(SMD), y = variable, 
      group = Method, color = Method)
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1,  #<<
             color = "black", size = 0.1) #<<
```
]

---

class: title title-1

# What if we changed the caliper?

.small[
```{r}
matched_smds <- matchit(satisfied_customer_service ~ income + married +  
                          n_kids + has_pets + age + education + gender + 
                          former_customer_service + previous_spend, 
                        data = customer_satisfaction, caliper = 0.01) %>%
  get_matches() %>%
  summarise( 
    across( 
      c(income, married, n_kids, has_pets, age, 
        education, gender, former_customer_service, previous_spend), 
      list(
        matched.calp.01 = ~smd(.x, satisfied_customer_service)$estimate
      )
    )
  )
```
]

```{r, echo = FALSE}
plot_df_all <- matched_smds %>% #<<
  pivot_longer( 
    everything(),
    values_to = "SMD", 
    names_to = c("variable", "Method"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  bind_rows(plot_df_all) %>%
  arrange(Method, abs(SMD)) %>%
  mutate(variable = fct_inorder(variable))
```

---

class: title title-1

# Love Plot

.left-code[
```{r, eval = FALSE}
ggplot(
  data = plot_df_all,
  aes(x = abs(SMD), y = variable, 
      group = Method, color = Method)
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1,  
             color = "black", size = 0.1) 
```
]

.right-plot[
```{r, echo = FALSE}
ggplot(
  data = plot_df_all,
  aes(x = abs(SMD), y = variable, 
      group = Method, color = Method)
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1,  #<<
             color = "black", size = 0.1) #<<
```
]

---
class: title title-1

# ECDF

For continuous variables, it can be helpful to look at the _whole_ distribution pre and post-weighting rather than a single summary measure

```{r, echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE}
ggplot(customer_satisfaction, aes(x = previous_spend, group = satisfied_customer_service, color = factor(satisfied_customer_service))) +
  stat_ecdf() +
  scale_color_manual("Satisfied with Customer Service", values = c("#5154B8", "#5DB854"),
                     labels = c("No", "Yes")) + 
  scale_x_continuous("Previous spending", label = scales::dollar) + 
  ylab("Proportion <= x") 
```

---

class: title title-1

# Unweighted ECDF

.small[
```{r, message = FALSE, warning = FALSE, eval = FALSE}
ggplot(customer_satisfaction, 
       aes(x = previous_spend, group = satisfied_customer_service, 
           color = factor(satisfied_customer_service))) +
  stat_ecdf() +
  scale_color_manual("Satisfied with Customer Service", 
                     values = c("#5154B8", "#5DB854"),
                     labels = c("No", "Yes")) + 
  scale_x_continuous("Previous spending", label = scales::dollar) + 
  ylab("Proportion <= x") 
```
]

---

class: title title-1

# Unweighted ECDF

.small[
```{r, message = FALSE, warning = FALSE, eval = FALSE}
ggplot(customer_satisfaction, #<<
       aes(x = previous_spend, group = satisfied_customer_service, #<<
           color = factor(satisfied_customer_service))) + #<<
  stat_ecdf() +
  scale_color_manual("Satisfied with Customer Service", 
                     values = c("#5154B8", "#5DB854"),
                     labels = c("No", "Yes")) + 
  scale_x_continuous("Previous spending", label = scales::dollar) + 
  ylab("Proportion <= x") 
```
]

---

class: title title-1

# Unweighted ECDF

.small[
```{r, message = FALSE, warning = FALSE, eval = FALSE}
ggplot(customer_satisfaction, 
       aes(x = previous_spend, group = satisfied_customer_service, 
           color = factor(satisfied_customer_service))) +
  stat_ecdf() + #<<
  scale_color_manual("Satisfied with Customer Service", 
                     values = c("#5154B8", "#5DB854"),
                     labels = c("No", "Yes")) + 
  scale_x_continuous("Previous spending", label = scales::dollar) + 
  ylab("Proportion <= x") 
```
]

---

class: title title-1

# Unweighted ECDF

.small[
```{r, message = FALSE, warning = FALSE, eval = FALSE}
ggplot(customer_satisfaction, 
       aes(x = previous_spend, group = satisfied_customer_service, 
           color = factor(satisfied_customer_service))) +
  stat_ecdf() +
  scale_color_manual("Satisfied with Customer Service", #<< 
                     values = c("#5154B8", "#5DB854"), #<<
                     labels = c("No", "Yes")) +  #<<
  scale_x_continuous("Previous spending", label = scales::dollar) + 
  ylab("Proportion <= x") 
```
]

---

class: title title-1

# Unweighted ECDF

.small[
```{r, message = FALSE, warning = FALSE, eval = FALSE}
ggplot(customer_satisfaction, 
       aes(x = previous_spend, group = satisfied_customer_service, 
           color = factor(satisfied_customer_service))) +
  stat_ecdf() +
  scale_color_manual("Satisfied with Customer Service", 
                     values = c("#5154B8", "#5DB854"),
                     labels = c("No", "Yes")) + 
  scale_x_continuous("Previous spending", label = scales::dollar) +  #<<
  ylab("Proportion <= x") 
```
]

---

class: title title-1

# Unweighted ECDF

.small[
```{r, message = FALSE, warning = FALSE, eval = FALSE}
ggplot(customer_satisfaction, 
       aes(x = previous_spend, group = satisfied_customer_service, 
           color = factor(satisfied_customer_service))) +
  stat_ecdf() +
  scale_color_manual("Satisfied with Customer Service", 
                     values = c("#5154B8", "#5DB854"),
                     labels = c("No", "Yes")) + 
  scale_x_continuous("Previous spending", label = scales::dollar) + 
  ylab("Proportion <= x")  #<<
```
]

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.box-1[Create an unweighted eCDF looking at `age`]

.box-1[Knit, commit, push to GitHub]

`r countdown::countdown(6)`

---
class: title title-1

# Weighted  ECDF 

.small[
```{r, echo = FALSE}
ecdf_1 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 1) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ecdf_0 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 0) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ggplot(ecdf_1, aes(x = previous_spend, y = cum_pct)) +
  geom_line(color = "#5DB854") +
  geom_line(data = ecdf_0, 
            aes(x = previous_spend, y = cum_pct), 
            color = "#5154B8") + 
  xlab("Previous spending") + 
  ylab("Proportion <= x") 
```
]

---

class: title title-1

# Weighted  ECDF 

.small[
```{r, eval = FALSE}
ecdf_1 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 1) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ecdf_0 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 0) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ggplot(ecdf_1, aes(x = previous_spend, y = cum_pct)) +
  geom_line(color = "#5DB854") +
  geom_line(data = ecdf_0, 
            aes(x = previous_spend, y = cum_pct), 
            color = "#5154B8") + 
  xlab("Previous spending") + 
  ylab("Proportion <= x") 
```
]
---

class: title title-1

# Weighted  ECDF 

.small[
```{r, eval = FALSE}
ecdf_1 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 1) %>% #<<
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ecdf_0 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 0) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ggplot(ecdf_1, aes(x = previous_spend, y = cum_pct)) +
  geom_line(color = "#5DB854") +
  geom_line(data = ecdf_0, 
            aes(x = previous_spend, y = cum_pct), 
            color = "#5154B8") + 
  xlab("Previous spending") + 
  ylab("Proportion <= x") 
```
]

---

class: title title-1

# Weighted  ECDF 

.small[
```{r, eval = FALSE}
ecdf_1 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 1) %>%
  arrange(previous_spend) %>% #<<
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ecdf_0 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 0) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ggplot(ecdf_1, aes(x = previous_spend, y = cum_pct)) +
  geom_line(color = "#5DB854") +
  geom_line(data = ecdf_0, 
            aes(x = previous_spend, y = cum_pct), 
            color = "#5154B8") + 
  xlab("Previous spending") + 
  ylab("Proportion <= x") 
```
]

---


class: title title-1

# Weighted  ECDF 

.small[
```{r, eval = FALSE}
ecdf_1 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 1) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts)) #<<
ecdf_0 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 0) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ggplot(ecdf_1, aes(x = previous_spend, y = cum_pct)) +
  geom_line(color = "#5DB854") +
  geom_line(data = ecdf_0, 
            aes(x = previous_spend, y = cum_pct),
            color = "#5154B8") + 
  xlab("Previous spending") + 
  ylab("Proportion <= x") 
```
]
---

class: title title-1

# Weighted  ECDF 
.small[
```{r, eval = FALSE}
ecdf_1 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 1) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ecdf_0 <- customer_satisfaction %>% #<<
  filter(satisfied_customer_service == 0) %>% #<<
  arrange(previous_spend) %>% #<<
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts)) #<<
ggplot(ecdf_1, aes(x = previous_spend, y = cum_pct)) +
  geom_line(color = "#5DB854") +
  geom_line(data = ecdf_0,
            aes(x = previous_spend, y = cum_pct), 
            color = "#5154B8") + 
  xlab("Previous spending") + 
  ylab("Proportion <= x") 
```
]
---
class: title title-1

# Weighted  ECDF 
.small[
```{r, eval = FALSE}
ecdf_1 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 1) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ecdf_0 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 0) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ggplot(ecdf_1, aes(x = previous_spend, y = cum_pct)) + #<<
  geom_line(color = "#5DB854") +
  geom_line(data = ecdf_0, 
            aes(x = previous_spend, y = cum_pct), 
            color = "#5154B8") + 
  xlab("Previous spending") + 
  ylab("Proportion <= x") 
```
]
---

class: title title-1

# Weighted  ECDF 

.small[
```{r, eval = FALSE}
ecdf_1 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 1) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ecdf_0 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 0) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ggplot(ecdf_1, aes(x = previous_spend, y = cum_pct)) +
  geom_line(color = "#5DB854") + #<<
  geom_line(data = ecdf_0, 
            aes(x = previous_spend, y = cum_pct), 
            color = "#5154B8") + 
  xlab("Previous spending") + 
  ylab("Proportion <= x") 
```
]
---

class: title title-1

# Weighted  ECDF 

.small[
```{r, eval = FALSE}
ecdf_1 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 1) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ecdf_0 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 0) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ggplot(ecdf_1, aes(x = previous_spend, y = cum_pct)) +
  geom_line(color = "#5DB854") +
  geom_line(data = ecdf_0, #<<
            aes(x = previous_spend, y = cum_pct), #<<
            color = "#5154B8") +  #<<
  xlab("Previous spending") + 
  ylab("Proportion <= x") 
```

]

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.box-1[Create an weighted eCDF looking at `age` with the `ate` weights]

.box-1[Create an weighted eCDF looking at `age` with the `ato` weights]

.box-1[Knit, commit, push to GitHub]

`r countdown::countdown(6)`

---

class: title title-1

# Iterate on your propensity score model!


.box-1.medium[Decrease your caliper if doing matching]

--

.box-1.medium[Allow for more degrees of freedom for variables that look imbalanced]

--

.box-inv-1.medium[Polynomial terms // splines]



---

class: title title-1

# Iterate on your propensity score model

.box-1[`previous_spend` modeled as a 3rd degree polynomial]

.small[
```{r}
glm(satisfied_customer_service ~ income + married + n_kids + has_pets + age + 
      education + gender + former_customer_service + 
      poly(previous_spend, 3), 
    family = binomial(),
    data = customer_satisfaction) %>%
  augment(data = customer_satisfaction, type.predict = "response") %>%
  mutate(atm_wts = pmin(.fitted, 1 - .fitted) / 
           (satisfied_customer_service * .fitted + 
              (1 - satisfied_customer_service) * (1 - .fitted))
  ) -> customer_satisfaction
```
]

---

class: title title-1

# Iterate on your propensity score model

.box-1[`previous_spend` modeled as a 3rd degree polynomial]

.small[
```{r, eval = FALSE}
glm(satisfied_customer_service ~ income + married + n_kids + has_pets + age + 
      education + gender + former_customer_service + 
      poly(previous_spend, 3), #<<
    family = binomial(),
    data = customer_satisfaction) %>%
  augment(data = customer_satisfaction, type.predict = "response") %>%
  mutate(atm_wts = pmin(.fitted, 1 - .fitted) / 
           (satisfied_customer_service * .fitted + 
              (1 - satisfied_customer_service) * (1 - .fitted))
  ) -> customer_satisfaction
```
]
---

class: title title-1

# Iterate on your propensity score model

.box-1.small[`previous_spend` modeled as a 3rd degree polynomial]

```{r, echo = FALSE, fig.height = 5}
customer_satisfaction %>%
  summarise(
    across( 
      c(income, married, n_kids, has_pets, age, 
        education, gender, former_customer_service, previous_spend), 
      list(
        unweighted = ~smd(.x, satisfied_customer_service)$estimate, 
        weighted = ~smd(.x, satisfied_customer_service, atm_wts)$estimate 
      )
    )
  ) %>%
  pivot_longer(
    everything(),
    values_to = "SMD", 
    names_to = c("variable", "Method"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  arrange(Method, abs(SMD)) %>%
  mutate(variable = fct_inorder(variable)) %>%
  ggplot(
  aes(x = abs(SMD), y = variable, 
      group = Method, color = Method)
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1,  
             color = "black", size = 0.1)
```

---

class: title title-1

# Iterate on your propensity score model

.box-1.small[`previous_spend` modeled as a 3rd degree polynomial]

```{r, echo = FALSE, fig.height = 5}
ecdf_1 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 1) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ecdf_0 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 0) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ggplot(ecdf_1, aes(x = previous_spend, y = cum_pct)) +
  geom_line(color = "#5DB854") +
  geom_line(data = ecdf_0, #<<
            aes(x = previous_spend, y = cum_pct), #<<
            color = "#5154B8") +  #<<
  xlab("Previous spending") + 
  ylab("Proportion <= x") 
```

---


class: title title-1

# Iterate on your propensity score model

.box-1[`previous_spend` modeled as a natural spline with 3 degrees of freedom]

.small[
```{r}
glm(satisfied_customer_service ~ income + married + n_kids + has_pets + age + 
      education + gender + former_customer_service + 
      splines::ns(previous_spend, 3),
    family = binomial(),
    data = customer_satisfaction) %>%
  augment(data = customer_satisfaction, type.predict = "response") %>%
  mutate(atm_wts = pmin(.fitted, 1 - .fitted) / 
           (satisfied_customer_service * .fitted + 
              (1 - satisfied_customer_service) * (1 - .fitted))
  ) -> customer_satisfaction
```
]
---

class: title title-1

# Iterate on your propensity score model

.box-1.small[`previous_spend` modeled as a natural spline with 3 degrees of freedom]

.small[
```{r, eval = FALSE}
glm(satisfied_customer_service ~ income + married + n_kids + has_pets + age + 
      education + gender + former_customer_service + 
      splines::ns(previous_spend, 3), #<<
    family = binomial(),
    data = customer_satisfaction) %>%
  augment(data = customer_satisfaction, type.predict = "response") %>%
  mutate(atm_wts = pmin(.fitted, 1 - .fitted) / 
           (satisfied_customer_service * .fitted + 
              (1 - satisfied_customer_service) * (1 - .fitted))
  ) -> customer_satisfaction
```
]
---

class: title title-1

# Iterate on your propensity score model

.box-1.small[`previous_spend` modeled as a natural spline with 3 degrees of freedom]

```{r, echo = FALSE, fig.height = 5}
customer_satisfaction %>%
  summarise(
    across( 
      c(income, married, n_kids, has_pets, age, 
        education, gender, former_customer_service, previous_spend), 
      list(
        unweighted = ~smd(.x, satisfied_customer_service)$estimate, 
        weighted = ~smd(.x, satisfied_customer_service, atm_wts)$estimate 
      )
    )
  ) %>%
  pivot_longer(
    everything(),
    values_to = "SMD", 
    names_to = c("variable", "Method"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  arrange(Method, abs(SMD)) %>%
  mutate(variable = fct_inorder(variable)) %>%
  ggplot(
  aes(x = abs(SMD), y = variable, 
      group = Method, color = Method)
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1,  
             color = "black", size = 0.1)
```

---

class: title title-1

# Iterate on your propensity score model

.box-1[`previous_spend` modeled as a natural spline with 3 degrees of freedom]

```{r, echo = FALSE, fig.height = 5}
ecdf_1 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 1) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ecdf_0 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 0) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ggplot(ecdf_1, aes(x = previous_spend, y = cum_pct)) +
  geom_line(color = "#5DB854") +
  geom_line(data = ecdf_0, #<<
            aes(x = previous_spend, y = cum_pct), #<<
            color = "#5154B8") +  #<<
  xlab("Previous spending") + 
  ylab("Proportion <= x") 
```


---



class: title title-1

# Iterate on your propensity score model

.box-1.small[`previous_spend` modeled as a natural spline with 8 degrees of freedom]

.small[
```{r}
glm(satisfied_customer_service ~ income + married + n_kids + has_pets + age + 
      education + gender + former_customer_service + 
      splines::ns(previous_spend, 8), #<<
    family = binomial(),
    data = customer_satisfaction) %>%
  augment(data = customer_satisfaction, type.predict = "response") %>%
  mutate(atm_wts = pmin(.fitted, 1 - .fitted) / 
           (satisfied_customer_service * .fitted + 
              (1 - satisfied_customer_service) * (1 - .fitted))
  ) -> customer_satisfaction
```
]
---

class: title title-1

# Iterate on your propensity score model

.box-1.small[`previous_spend` modeled as a natural spline with 8 degrees of freedom]

```{r, echo = FALSE, fig.height = 5}
customer_satisfaction %>%
  summarise(
    across( 
      c(income, married, n_kids, has_pets, age, 
        education, gender, former_customer_service, previous_spend), 
      list(
        unweighted = ~smd(.x, satisfied_customer_service)$estimate, 
        weighted = ~smd(.x, satisfied_customer_service, atm_wts)$estimate 
      )
    )
  ) %>%
  pivot_longer(
    everything(),
    values_to = "SMD", 
    names_to = c("variable", "Method"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  arrange(Method, abs(SMD)) %>%
  mutate(variable = fct_inorder(variable)) %>%
  ggplot(
  aes(x = abs(SMD), y = variable, 
      group = Method, color = Method)
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1,  
             color = "black", size = 0.1)
```

---

class: title title-1

# Iterate on your propensity score model

.box-1[`previous_spend` modeled as a natural spline with 8 degrees of freedom]

```{r, echo = FALSE, fig.height = 5}
ecdf_1 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 1) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ecdf_0 <- customer_satisfaction %>%
  filter(satisfied_customer_service == 0) %>%
  arrange(previous_spend) %>%
  mutate(cum_pct = cumsum(atm_wts) / sum(atm_wts))
ggplot(ecdf_1, aes(x = previous_spend, y = cum_pct)) +
  geom_line(color = "#5DB854") +
  geom_line(data = ecdf_0, #<<
            aes(x = previous_spend, y = cum_pct), #<<
            color = "#5154B8") +  #<<
  xlab("Previous spending") + 
  ylab("Proportion <= x") 
```

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.box-1[Update your propensity score model to include at least one natural spline]

.box-1[Recreate your Love Plot and eCDF plots to examine how (whether) this impacts your results]

.box-1[Knit, commit, push to GitHub]

`r countdown::countdown(10)`

