---
title: "Estimating Causal Effects"
author: "Lucy D'Agostino McGowan"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    css: ["default", "css/style.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```


```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view"))
```

class: center middle main-title section-title-1

# Estimating Causal Effects

.class-info[

**Session 15**

.light[STA 379/679: Causal Inference <br>
Lucy D'Agostino McGowan
]

]

---

class: title title-1

# Weighted Estimator

.box-1[
$$\hat{\tau} = \frac{\sum_{i=1}^n Y_i X_i w_i}{\sum_{i=1}^nX_iw_i}-\frac{\sum_{i=1}^n Y_i (1-X_i) w_i}{\sum_{i=1}^n(1-X_i)w_i}$$
]

---
class: title title-1

# Weighted Estimator

.box-1[

$$\hat{\tau}_{ATE} = \frac{1}{n}\sum_{i=1}^n\left\{\frac{X_iY_i}{\hat{e}(\mathbf{Z_i})}-\frac{(1-X_i)Y_i}{1-\hat{e}(\mathbf{Z_i})}\right\}$$
]

--
<br>

.box-1[
Is this an unbiased estimator for $\tau_{ATE}$?
]

---

class: title title-1

# Using the propensity score
.footer[Imbens and Rubin (2015) Causal Inference]

.box-1[We can show that]

.box-inv-1[

$$E\left[\frac{Y_i^{obs}X_i}{e(Z_i)}\right] = E[Y_i(1)]$$

]

--
<br>

.box-inv-1[

$$E\left[\frac{Y_i^{obs}(1 - X_i)}{1 - e(Z_i)}\right] = E[Y_i(0)]$$

]

---

class: title title-1

# Using the propensity score
.footer[Imbens and Rubin (2015) Causal Inference]

.box-1[

$$E\left[\frac{Y_i^{obs}X_i}{e(Z_i)}\right]=E\left[E\left[\frac{Y_i^{obs}X_i}{e(Z_i)}\Big|Z_i\right]\right]$$
]

--

.box-inv-1.medium[Why?]

--

.box-1[Law of iterated expectations]

---

class: title title-1

# Using the propensity score
.footer[Imbens and Rubin (2015) Causal Inference]

.box-1[

$$E\left[E\left[\frac{Y_i^{obs}X_i}{e(Z_i)}\Big|Z_i\right]\right] = E\left[E\left[\frac{Y_i(1)X_i}{e(Z_i)}\Big|Z_i\right]\right]$$
]

---

class: title title-1

# Using the propensity score
.footer[Imbens and Rubin (2015) Causal Inference]

.box-1[

$$E\left[E\left[\frac{\color{orange}{Y_i^{obs}}X_i}{e(Z_i)}\Big|Z_i\right]\right] = E\left[E\left[\frac{\color{orange}{Y_i(1)}X_i}{e(Z_i)}\Big|Z_i\right]\right]$$
]

--

.box-inv-1.medium[Why?]

--

.box-1[We're only looking at the observations where $X_i = 1$]

---
class: title title-1

# Using the propensity score
.footer[Imbens and Rubin (2015) Causal Inference]

.box-1[

$$=E\left[E\left[\frac{Y_i(1)X_i}{e(Z_i)}\Big|Z_i\right]\right]=E\left[\frac{E[Y_i(1)|Z_i]E_X[X_i|Z_i]}{e(Z_i)}\right]$$
]

---
class: title title-1

# Using the propensity score
.footer[Imbens and Rubin (2015) Causal Inference]

.box-1[

$$=E\left[E\left[\frac{\color{orange}{Y_i(1)X_i}}{e(Z_i)}\Big|\color{orange}{Z_i}\right]\right]=E\left[\frac{\color{orange}{E[Y_i(1)|Z_i]E_X[X_i|Z_i]}}{e(Z_i)}\right]$$
]

--

.box-inv-1.medium[Why?]

--

.box-1[
Unconfoundedness (we can split $E[Y(1)X]=E[Y(1)]E[X]$)
]

---
class: title title-1

# Using the propensity score
.footer[Imbens and Rubin (2015) Causal Inference]

.box-1[

$$E\left[\frac{E[Y_i(1)|Z_i]E_X[X_i|Z_i]}{e(Z_i)}\right]=E[E[Y_i(1)|Z_i]]$$
]
---
class: title title-1

# Using the propensity score
.footer[Imbens and Rubin (2015) Causal Inference]

.box-1[

$$E\left[\frac{E[Y_i(1)|Z_i]\color{orange}{E_X[X_i|Z_i]}}{\color{orange}{e(Z_i)}}\right]=E[E[Y_i(1)|Z_i]]$$
]

--

.box-inv-1.medium[Why?]

--

.box-1[
propensity score definition: $e(Z_i)=E_X[X_i|Z_i]$
]

---
class: title title-1

# Using the propensity score
.footer[Imbens and Rubin (2015) Causal Inference]

.box-1[

$$E[E[Y_i(1)|Z_i]]=E[Y_i(1)]$$
]

--

.box-inv-1.medium[Why?]

--

.box-1[
Law of iterated expectation
]

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

.box-1[Open `appex-10`]

.box-1[Calculate the ATE and ATO estimates for the effect of `satisfied_customer_service` on `next_spend`]

.box-1[Knit, commit, push to GitHub]

`r countdown::countdown(10)`

