---
title: "Outcome Models"
author: "Lucy D'Agostino McGowan"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    css: ["default", "css/style.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```


```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view"))
```

class: center middle main-title section-title-1

# Outcome Models

.class-info[

**Session 16**

.light[STA 379/679: Causal Inference <br>
Lucy D'Agostino McGowan
]

]

---

class: title title-1

# Weighted Estimator

.box-1[
$$\hat{\tau} = \frac{\sum_{i=1}^n Y_i X_i w_i}{\sum_{i=1}^nX_iw_i}-\frac{\sum_{i=1}^n Y_i (1-X_i) w_i}{\sum_{i=1}^n(1-X_i)w_i}$$
]

--

.box-inv-1[
What about the variability?
]
---
class: title title-1

# Example

.small[
```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
coffee_town_df <- read_csv("../lab-03-using-propensity-scores/coffee_town_df.csv")
coffee_town_df
```
]


---

class: title title-1

# Example

.box-1.medium[Causal Question]

--

.box-inv-1[What is the average causal effect of living in coffee town on the number of cups of coffee consumed across the whole population?]

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

1. Calculate a propensity score model estimating for the exposure `coffee_town`
2. Calculate `ate` weights using the propensity score
3. Knit commit & push to GitHub

`r countdown::countdown(8, font_size = "2em")`

---

class: title title-1

# Example

.small[
```{r, echo = TRUE, message = FALSE, warning = FALSE}
library(broom)

coffee_town_df <- 
  glm(coffee_town ~ age + job + smoke, 
      data = coffee_town_df,
      family = binomial()) %>%
  augment(data = coffee_town_df, type.predict = "response") %>%
  mutate(
    ate = coffee_town / .fitted + (1 - coffee_town)  / (1 - .fitted)
  )
```
]

---

class: title title-1

# Estimating the ATE

.pull-left[
.small[
```{r}
coffee_town_df %>%
  summarise(
    ate = sum(cups_of_coffee * coffee_town * ate) / 
      sum(coffee_town * ate) - 
      sum(cups_of_coffee * (1 - coffee_town) * ate) / 
      sum((1 - coffee_town) * ate))
```
]
]

---


class: title title-1

# Estimating the ATE

.pull-left[
.small[
```{r}
coffee_town_df %>%
  summarise(
    ate = sum(cups_of_coffee * coffee_town * ate) / #<<
      sum(coffee_town * ate) - 
      sum(cups_of_coffee * (1 - coffee_town) * ate) / 
      sum((1 - coffee_town) * ate))
```
]
]

---


class: title title-1

# Estimating the ATE

.pull-left[
.small[
```{r}
coffee_town_df %>%
  summarise(
    ate = sum(cups_of_coffee * coffee_town * ate) / 
      sum(coffee_town * ate) - #<<
      sum(cups_of_coffee * (1 - coffee_town) * ate) / 
      sum((1 - coffee_town) * ate))
```
]
]

---


class: title title-1

# Estimating the ATE

.pull-left[
.small[
```{r}
coffee_town_df %>%
  summarise(
    ate = sum(cups_of_coffee * coffee_town * ate) / 
      sum(coffee_town * ate) - 
      sum(cups_of_coffee * (1 - coffee_town) * ate) /  #<<
      sum((1 - coffee_town) * ate))
```
]
]

---


class: title title-1

# Estimating the ATE

.pull-left[
.small[
```{r}
coffee_town_df %>%
  summarise(
    ate = sum(cups_of_coffee * coffee_town * ate) / 
      sum(coffee_town * ate) - 
      sum(cups_of_coffee * (1 - coffee_town) * ate) / 
      sum((1 - coffee_town) * ate)) #<<
```
]
]

---


class: title title-1

# Estimating the ATE

.pull-left[
.small[
```{r}
coffee_town_df %>%
  summarise(
    ate = sum(cups_of_coffee * coffee_town * ate) / 
      sum(coffee_town * ate) - 
      sum(cups_of_coffee * (1 - coffee_town) * ate) / 
      sum((1 - coffee_town) * ate))
```
]
]

.pull-right[
.small[
```{r}
lm(cups_of_coffee ~ coffee_town, 
   weights = ate, 
   data = coffee_town_df)
```
]
]

---
class: subtitle subtitle-1 middl

# Warning: This may not be the right estimate! We haven't checked / iterated, this is just for example purposes

---
class: title-1 title

# Estimating the ATE

# Estimating the ATE

.pull-left[
.small[
```{r}
coffee_town_df %>%
  summarise(
    ate = sum(cups_of_coffee * coffee_town * ate) / 
      sum(coffee_town * ate) - 
      sum(cups_of_coffee * (1 - coffee_town) * ate) / 
      sum((1 - coffee_town) * ate))
```
]
]

.pull-right[
.small[
```{r}
lm(cups_of_coffee ~ coffee_town, 
   weights = ate, #<<
   data = coffee_town_df)
```
]
]
---

class: title title-1

# Can we just use the CIs from `lm`?

.small[
```{r}
lm(cups_of_coffee ~ coffee_town, 
   weights = ate, #<<
   data = coffee_town_df) %>%
  tidy(conf.int = TRUE)
```
]

--

.box-1[No!]

---


class: title title-1

# Option 1: Bootstrap

.small[
```{r, eval = FALSE}
fit_ipw <- function(split, ...) {
  .df <- analysis(split)
  # fit propensity score model
  propensity_model <- glm(
    exposure ~ confounder_1 + confounder_2 + ...
    family = binomial(), 
    data = .df
  )
  # calculate inverse probability weights
  .df <- propensity_model %>% 
    augment(type.predict = "response", data = .df) %>% 
    mutate(wts = exposure / .fitted + (1 - exposure) / (1  - .fitted)) 
  # fit correctly bootsrapped ipw model
  lm(outcome ~ exposure, data = .df, weights = wts) %>% 
    tidy()
}
```
]
---

class: title title-1

# Option 1: Bootstrap

.small[
```{r, eval = FALSE}
fit_ipw <- function(split, ...) { #<<
  .df <- analysis(split)
  # fit propensity score model
  propensity_model <- glm(
    exposure ~ confounder_1 + confounder_2 + ...
    family = binomial(), 
    data = .df
  )
  # calculate inverse probability weights
  .df <- propensity_model %>% 
    augment(type.predict = "response", data = .df) %>% 
    mutate(wts = exposure / .fitted + (1 - exposure) / (1  - .fitted)) 
  # fit correctly bootsrapped ipw model
  lm(outcome ~ exposure, data = .df, weights = wts) %>% 
    tidy()
} #<<
```
]

---


class: title title-1

# Option 1: Bootstrap

.small[
```{r, eval = FALSE}
fit_ipw <- function(split, ...) {
  .df <- analysis(split) #<<
  # fit propensity score model
  propensity_model <- glm(
    exposure ~ confounder_1 + confounder_2 + ...
    family = binomial(), 
    data = .df
  )
  # calculate inverse probability weights
  .df <- propensity_model %>% 
    augment(type.predict = "response", data = .df) %>% 
    mutate(wts = exposure / .fitted + (1 - exposure) / (1  - .fitted)) 
  # fit correctly bootsrapped ipw model
  lm(outcome ~ exposure, data = .df, weights = wts) %>% 
    tidy()
}
```
]

---


class: title title-1

# Option 1: Bootstrap

.small[
```{r, eval = FALSE}
fit_ipw <- function(split, ...) {
  .df <- analysis(split)
  # fit propensity score model #<<
  propensity_model <- glm( #<<
    exposure ~ confounder_1 + confounder_2 + ... #<<
    family = binomial(),  #<<
    data = .df #<<
  ) #<<
  # calculate inverse probability weights
  .df <- propensity_model %>% 
    augment(type.predict = "response", data = .df) %>% 
    mutate(wts = exposure / .fitted + (1 - exposure) / (1  - .fitted)) 
  # fit correctly bootsrapped ipw model
  lm(outcome ~ exposure, data = .df, weights = wts) %>% 
    tidy()
}
```
]

---

class: title title-1

# Option 1: Bootstrap

.small[
```{r, eval = FALSE}
fit_ipw <- function(split, ...) {
  .df <- analysis(split)
  # fit propensity score model
  propensity_model <- glm(
    exposure ~ confounder_1 + confounder_2 + ...
    family = binomial(), 
    data = .df
  )
  # calculate inverse probability weights #<<
  .df <- propensity_model %>%  #<<
    augment(type.predict = "response", data = .df) %>%  #<<
    mutate(wts = exposure / .fitted + (1 - exposure) / (1  - .fitted)) #<<
  # fit correctly bootsrapped ipw model
  lm(outcome ~ exposure, data = .df, weights = wts) %>% 
    tidy()
}
```
]

---


class: title title-1

# Option 1: Bootstrap

.small[
```{r, eval = FALSE}
fit_ipw <- function(split, ...) {
  .df <- analysis(split)
  # fit propensity score model
  propensity_model <- glm(
    exposure ~ confounder_1 + confounder_2 + ...
    family = binomial(), 
    data = .df
  )
  # calculate inverse probability weights
  .df <- propensity_model %>% 
    augment(type.predict = "response", data = .df) %>% 
    mutate(wts = exposure / .fitted + (1 - exposure) / (1  - .fitted)) 
  # fit correctly bootsrapped ipw model #<<
  lm(outcome ~ exposure, data = .df, weights = wts) %>%  #<<
    tidy() #<<
}
```
]

---

class: title title-1

# Option 1: Bootstrap

.small[
```{r, eval = FALSE}
library(rsample)
# fit ipw model to bootstrapped samples
ipw_results <- bootstraps(df, 1000, apparent = TRUE) %>% 
  mutate(results = map(splits, fit_ipw))
```
]
---

class: title title-1

# Option 1: Bootstrap

.small[
```{r, eval = FALSE}
library(rsample) #<<
# fit ipw model to bootstrapped samples
ipw_results <- bootstraps(df, 1000, apparent = TRUE) %>% 
  mutate(results = map(splits, fit_ipw))
```
]

---

class: title title-1

# Option 1: Bootstrap

.small[
```{r, eval = FALSE}
library(rsample)
# fit ipw model to bootstrapped samples
ipw_results <- bootstraps(df, 1000, apparent = TRUE) %>% #<<
  mutate(results = map(splits, fit_ipw))
```
]

---

class: title title-1

# Option 1: Bootstrap

.small[
```{r, eval = FALSE}
library(rsample)
# fit ipw model to bootstrapped samples
ipw_results <- bootstraps(df, 1000, apparent = TRUE) %>% 
  mutate(results = map(splits, fit_ipw)) #<<
```
]

---

class: title title-1

# Option 1: Bootstrap

```{r, eval = FALSE}
# get t-statistic-based CIs
boot_estimate <- int_t(ipw_results, results) %>%
  filter(term == "exposure")
```

---

class: title title-1

# Option 1: Bootstrap

```{r, eval = FALSE}
# get t-statistic-based CIs
boot_estimate <- int_t(ipw_results, results) %>% #<<
  filter(term == "exposure")
```

---

class: title title-1

# Option 1: Bootstrap

```{r, eval = FALSE}
# get t-statistic-based CIs
boot_estimate <- int_t(ipw_results, results) %>%
  filter(term == "exposure") #<<
```

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

1. `install.packages("rsample")`
2. Create a function called that fits the propensity score model and the weighted outcome model for the effect between `coffee_town` and `cups_of_coffee`
3. Using the `bootstraps()` and `int_t()` functions to estimate the final effect.
4. Knit commit & push to GitHub

`r countdown::countdown(8, font_size = "2em")`

---

class: title title-1

# Bootstrap estimate

```{r, echo = FALSE, cache = TRUE}
fit_ipw <- function(split) {
  .df <- analysis(split)
  
  # fit propensity score model
  propensity_model <- glm(
    coffee_town ~ age + job + smoke,
    family = binomial(),
    data = .df
  )
  
  # calculate inverse probability weights
  .df <- propensity_model %>%
    augment(type.predict = "response", data = .df) %>%
    mutate(wts = coffee_town / .fitted + (1 - coffee_town) / (1  - .fitted))
  
  # fit correctly bootsrapped ipw model
  lm(cups_of_coffee ~ coffee_town, data = .df, weights = wts) %>%
    tidy()
}


library(rsample)
# fit ipw model to bootstrapped samples
ipw_results <- bootstraps(coffee_town_df, 1000, apparent = TRUE) %>% 
  mutate(results = map(splits, fit_ipw))
boot_estimate <- int_t(ipw_results, results) %>%
  filter(term == "coffee_town")
boot_estimate
```

--

.box-1.medium[Naive estimate]

```{r, echo = FALSE}
lm(cups_of_coffee ~ coffee_town, data = coffee_town_df, weights = ate) %>%
  tidy(conf.int = TRUE) %>%
  select(term, estimate, conf.low, conf.high)
```

---

class: title title-1

# Option 2: Robust Standard Errors

```{r, eval = FALSE}
library(survey)
des <- svydesign(
  ids = ~1, 
  data = coffee_town_df,
  weights = ~ate
)
svyglm(cups_of_coffee ~ coffee_town, des) %>%
  tidy(conf.int = TRUE)
```
---

class: title title-1

# Option 2: Robust Standard Errors

```{r, eval = FALSE}
library(survey) #<<
des <- svydesign(
  ids = ~1, 
  data = coffee_town_df,
  weights = ~ate
)
svyglm(cups_of_coffee ~ coffee_town, des) %>%
  tidy(conf.int = TRUE)
```

---
class: title title-1

# Option 2: Robust Standard Errors

```{r, eval = FALSE}
library(survey)
des <- svydesign( #<<
  ids = ~1,  #<<
  data = coffee_town_df, #<<
  weights = ~ate #<<
) #<<
svyglm(cups_of_coffee ~ coffee_town, des) %>%
  tidy(conf.int = TRUE)
```

---

class: title title-1

# Option 2: Robust Standard Errors

```{r, eval = FALSE}
library(survey)
des <- svydesign(
  ids = ~1, 
  data = coffee_town_df,
  weights = ~ate
)
svyglm(cups_of_coffee ~ coffee_town, des) %>% #<<
  tidy(conf.int = TRUE)
```

---



class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

1. Use the `svyglm` function from the survey package to model the average treatment effect with the robust standard errors.
2. Knit commit & push to GitHub

`r countdown::countdown(5, font_size = "2em")`

---

class: title title-1

# Option 2: Robust Standard Errors

```{r, echo = FALSE}
library(survey)
des <- svydesign(
  ids = ~1, 
  data = coffee_town_df,
  weights = ~ate
)
svyglm(cups_of_coffee ~ coffee_town, des) %>%
  tidy(conf.int = TRUE) %>%
  select(term, estimate, conf.low, conf.high)
```

---
class: title title-1

# Option 3: Sandwich estimator

.box-1[(that takes the propensity score estimation into account)]

--

.box-1[
$$u(\theta)=\begin{pmatrix}(Y-\mu_1)Xe(\mathbf{Z}, \beta)^{-1}\\(Y-\mu_0)(1-X)(1-e(\mathbf{Z}, \beta))^{-1}\\\mathbf{Z}(X-e(\mathbf{Z}, \beta))\end{pmatrix}$$
]

--
<br>
.box-1[
$e(\mathbf{Z}, \beta)$ = the propensity score
]

---

class: title title-1

# Option 3: Sandwich estimator

.box-1[(that takes the propensity score estimation into account)]


.box-1[
$$u(\theta)=\begin{pmatrix}(Y-\mu_1)\color{orange}{Xe(\mathbf{Z}, \beta)^{-1}}\\(Y-\mu_0)\color{orange}{(1-X)(1-e(\mathbf{Z}, \beta))^{-1}}\\\mathbf{Z}(X-e(\mathbf{Z}, \beta))\end{pmatrix}$$
]

--

<br>

.box-1[ATE Weight]

---
class: title title-1

# Option 3: Sandwich estimator

.box-1[(that takes the propensity score estimation into account)]



.box-1[
$$u(\theta)=\begin{pmatrix}(Y-\mu_1)Xe(\mathbf{Z}, \beta)^{-1}\\(Y-\mu_0)(1-X)(1-e(\mathbf{Z}, \beta))^{-1}\\\color{orange}{\mathbf{Z}(X-e(\mathbf{Z}, \beta))}\end{pmatrix}$$
]

--
<br>
.box-1[
Estimating equation for the propensity score model
]


---

class: title title-1

# M-Estimation

.box-1[
$$\sqrt{n}(\hat\theta-\theta)\rightarrow^d N(0, \mathbf{A}^{-1}\mathbf{B}\mathbf{A}^{-T})$$
]

---

class: title title-1

# Sandwich estimator

.box-1[

$$\textrm{var}(\theta) = \mathbf{A}^{-1}\mathbf{B}\mathbf{A}^{-T}$$
]

--

<br>

.box-1[

$$\mathbf{A} = -E\left[\frac{\partial \mathbf{u}(\theta)}{\partial\theta}\right]$$
]
---

class: title title-1

# Sandwich estimator

.box-1[

$$\textrm{var}(\theta) = \mathbf{A}^{-1}\mathbf{B}\mathbf{A}^{-T}$$
]

--

<br>

.box-1[

$$\mathbf{B} = E[\mathbf{u}(\theta)\mathbf{u}(\theta)^T]$$
]

---

class: title title-1

# Estimating **A**

.box-1[
$$\mathbf{A} = -E\left[\frac{\partial \mathbf{u}(\theta)}{\partial\theta}\right]$$
]

--
<br>

.box-1[

$$\mathbf{A} = \begin{pmatrix}a_{11}& 0& a_{13}\\ 0 & a_{22} & a_{23}\\0&0&a_{33}\end{pmatrix}$$
]

---

class: title title-1

# Estimating **A**

.box-1[
$a_{11} = E[X e(\mathbf{Z}, \beta)^{-1}]$
]

--
<br>
.box-1[Why?]

.box-1[
$$-E\left[\frac{\partial(Y-\mu_1)Xe(\mathbf{Z}, \beta)^{-1}}{\partial \mu_1}\right]\\=-E[-Xe(\mathbf{Z}, \beta)^{-1}]$$
]
---

class: title title-1

# Estimating **A**

.box-1[
$a_{11} = E[X e(\mathbf{Z}, \beta)^{-1}]$
]

--
<br>
.box-1[
$a_{22} = E[(1-X) (1 - e(\mathbf{Z}, \beta))^{-1}]$
]

--
<br>
.box-1[
$a_{33} = E[\mathbf{ZZ}^Te(\mathbf{Z}, \beta)(1-e(\mathbf{Z}, \beta))]$

]

---

class: title title-1

# Estimating **A**

.box-1[
$a_{13} = E[\mathbf{Z}^T(Y-\mu_1)Xe(\mathbf{Z}, \beta)^{-1}(1-e(\mathbf{X}, \beta))]$
]

--

<br>
.box-1[
$a_{23} = E[\mathbf{ZZ}^T(X-e(\mathbf{Z}, \beta))^2]$
]

---

class: title title-1

# Estimating **A**

.box-1[
Condition on $\mathbf{Z}$ and knowing $E[X | \mathbf{Z}] = e(\mathbf{Z}, \beta)$]

--

.box-1[
$a_{11} = E[X e(\mathbf{Z}, \beta)^{-1}] = 1$
]

--

.box-1[Why?]

`r countdown::countdown(4, font_size = "2em")`

---

class: title title-1

# Estimating **A**

.box-1[
Condition on $\mathbf{Z}$ and knowing $E[X | \mathbf{Z}] = e(\mathbf{Z}, \beta)$]

--

.box-1[
$a_{22} = E[(1-X) (1 - e(\mathbf{Z}, \beta))^{-1}] = ?$
]


`r countdown::countdown(4, font_size = "2em")`

---

class: title title-1

# Estimating **A**

.box-1[
Condition on $\mathbf{Z}$ and knowing $E[X | \mathbf{Z}] = e(\mathbf{Z}, \beta)$]

--

.box-1[
$a_{22} = E[(1-X) (1 - e(\mathbf{Z}, \beta))^{-1}] = ?$
]


`r countdown::countdown(4, font_size = "2em")`

---

class: title title-1

# Estimating **A**

.box-1[
Condition on $\mathbf{Z}$ and knowing $E[X | \mathbf{Z}] = e(\mathbf{Z}, \beta)$]

<br>
.box-1[
$\hat{a}_{33} = \mathbf{ZZ}^T\hat{e}(1-\hat{e})$

]

---

class: title title-1

# Estimating **B**

.box-1[

$$\mathbf{B} = E[\mathbf{u}(\theta)\mathbf{u}(\theta)^T]$$
]

--
<br>
.box-1[
$$\mathbf{B}=\begin{pmatrix}b_{11}&0&b_{13}\\0&b_{22}&b_{23}\\b_{13}^T&b_{23}^T& b_{33}\end{pmatrix}$$
]

---
class: title-1 title
# Estimating **B**

.box-1[
$b_{11}=E[(Y_1-\mu_1)^2e^{-1}]$
]

--

<br>

.box-1[
$b_{13}=E[\mathbf{Z}^T(Y_1-\mu_1)(1-e)]$
]

---

class: title-1 title
# Estimating **B**

.box-1[
$b_{22}=E[(Y_0-\mu_0)^2(1-e)^{-1}]$
]

--

<br>

.box-1[
$b_{23}=-E[\mathbf{Z}^T(Y_0-\mu_0)e]$
]

--

<br>
.box-1[
$b_{33}=E[\mathbf{ZZ}^Te(1-e)]$
]

---

class: title title-1

# Putting the variance together

.box-1[
$\textrm{var}(\hat\mu_1)=n^{-1}(b_{11}-b_{13}b_{33}^{-1}b_{13}^T)$
]

--

<br>

.box-1[
$\textrm{var}(\hat\mu_0)=n^{-1}(b_{22}-b_{23}b_{33}^{-1}b_{23}^T)$
]

--

<br>

.box-1[
$\textrm{cov}(\hat\mu_1, \hat\mu_0)=-n^{-1}b_{13}b_{33}^{-1}b_{23}^T$
]

---

class: title title-1

# Doing it in R!

.box-1[
We need a bit of data manipulation
]
```{r, eval = FALSE}
library(PSW) 
coffee_town_df <- coffee_town_df %>%
  mutate(job_hard = ifelse(job == "hard", 1, 0),
         job_easy = ifelse(job == "easy", 1, 0),
         smoke_former = ifelse(smoke == "former", 1, 0),
         smoke_never = ifelse(smoke == "never", 1, 0)
  ) %>%
  as.data.frame()

```

---

class: title title-1

# Doing it in R!

.box-1[
We need a bit of data manipulation
]
```{r, eval = FALSE}
library(PSW)  #<<
coffee_town_df <- coffee_town_df %>%
  mutate(job_hard = ifelse(job == "hard", 1, 0),
         job_easy = ifelse(job == "easy", 1, 0),
         smoke_former = ifelse(smoke == "former", 1, 0),
         smoke_never = ifelse(smoke == "never", 1, 0)
  ) %>%
  as.data.frame()

```

---

class: title title-1

# Doing it in R!

.box-1[
We need a bit of data manipulation
]
```{r, eval = FALSE}
library(PSW) 
coffee_town_df <- coffee_town_df %>%
  mutate(job_hard = ifelse(job == "hard", 1, 0), #<<
         job_easy = ifelse(job == "easy", 1, 0),
         smoke_former = ifelse(smoke == "former", 1, 0),
         smoke_never = ifelse(smoke == "never", 1, 0)
  ) %>%
  as.data.frame()

```

---
class: title title-1

# Doing it in R!

.box-1[
We need a bit of data manipulation
]
```{r, eval = FALSE}
library(PSW) 
coffee_town_df <- coffee_town_df %>%
  mutate(job_hard = ifelse(job == "hard", 1, 0),
         job_easy = ifelse(job == "easy", 1, 0), #<<
         smoke_former = ifelse(smoke == "former", 1, 0),
         smoke_never = ifelse(smoke == "never", 1, 0)
  ) %>%
  as.data.frame()

```

---class: title title-1

# Doing it in R!

.box-1[
We need a bit of data manipulation
]
```{r, eval = FALSE}
library(PSW) 
coffee_town_df <- coffee_town_df %>%
  mutate(job_hard = ifelse(job == "hard", 1, 0),
         job_easy = ifelse(job == "easy", 1, 0),
         smoke_former = ifelse(smoke == "former", 1, 0), #<<
         smoke_never = ifelse(smoke == "never", 1, 0)
  ) %>%
  as.data.frame()

```

---
class: title title-1

# Doing it in R!

.box-1[
We need a bit of data manipulation
]
```{r, eval = FALSE}
library(PSW) 
coffee_town_df <- coffee_town_df %>%
  mutate(job_hard = ifelse(job == "hard", 1, 0),
         job_easy = ifelse(job == "easy", 1, 0),
         smoke_former = ifelse(smoke == "former", 1, 0),
         smoke_never = ifelse(smoke == "never", 1, 0) #<<
  ) %>%
  as.data.frame()

```

---
class: title title-1

# Doing it in R!

.box-1[
We need a bit of data manipulation
]
```{r}
library(PSW) 
coffee_town_df <- coffee_town_df %>%
  mutate(job_hard = ifelse(job == "hard", 1, 0),
         job_easy = ifelse(job == "easy", 1, 0),
         smoke_former = ifelse(smoke == "former", 1, 0),
         smoke_never = ifelse(smoke == "never", 1, 0)
  ) %>%
  as.data.frame() #<<

```


---

class: title title-1

# Doing it in R!
.small[
```{r}
mod <- psw(data = coffee_town_df, 
    form.ps = 
      "coffee_town ~ age + job_hard + job_easy +
       smoke_former + smoke_never",
    weight = "ATE",
    wt = TRUE,
    out.var = "cups_of_coffee", 
    family = "gaussian")
```

]


---


class: title title-1

# Doing it in R!
.small[
```{r}
mod <- psw(data = coffee_town_df, #<<
    form.ps = 
      "coffee_town ~ age + job_hard + job_easy +
       smoke_former + smoke_never",
    weight = "ATE",
    wt = TRUE,
    out.var = "cups_of_coffee", 
    family = "gaussian")
```

]
---


class: title title-1

# Doing it in R!
.small[
```{r}
mod <- psw(data = coffee_town_df, 
    form.ps =  #<<
      "coffee_town ~ age + job_hard + job_easy +#<<
       smoke_former + smoke_never",#<<
    weight = "ATE",
    wt = TRUE,
    out.var = "cups_of_coffee", 
    family = "gaussian")
```

]

---


class: title title-1

# Doing it in R!
.small[
```{r}
mod <- psw(data = coffee_town_df, 
    form.ps = 
      "coffee_town ~ age + job_hard + job_easy +
       smoke_former + smoke_never",
    weight = "ATE",#<<
    wt = TRUE,
    out.var = "cups_of_coffee", 
    family = "gaussian")
```

]
---


class: title title-1

# Doing it in R!
.small[
```{r}
mod <- psw(data = coffee_town_df, 
    form.ps = 
      "coffee_town ~ age + job_hard + job_easy +
       smoke_former + smoke_never",
    weight = "ATE",
    wt = TRUE, #<<
    out.var = "cups_of_coffee", 
    family = "gaussian")
```

]

---


class: title title-1

# Doing it in R!
.small[
```{r}
mod <- psw(data = coffee_town_df, 
    form.ps = 
      "coffee_town ~ age + job_hard + job_easy +
       smoke_former + smoke_never",
    weight = "ATE",
    wt = TRUE,
    out.var = "cups_of_coffee",  #<<
    family = "gaussian")
```

]

---


class: title title-1

# Doing it in R!
.small[
```{r}
mod <- psw(data = coffee_town_df, 
    form.ps = 
      "coffee_town ~ age + job_hard + job_easy +
       smoke_former + smoke_never",
    weight = "ATE",
    wt = TRUE,
    out.var = "cups_of_coffee", 
    family = "gaussian") #<<
```

]


---

class: title title-1

# Doing it in R!
.small[
```{r}
mod$est.wt
# lb
mod$est.wt - 1.96 * mod$std.wt

# ub
mod$est.wt + 1.96 * mod$std.wt
```
]

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

1. `install.packages("PSW")`
2. Calculate the large sample variance sandwich estimator using the `psw` function
3. Knit commit & push to GitHub

`r countdown::countdown(10, font_size = "2em")`

