---
title: "Sandwich Estimators"
author: "Lucy D'Agostino McGowan"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    css: ["default", "css/style.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```


```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view"))
```

class: center middle main-title section-title-1

# Sandwich Estimators

.class-info[

**Session 17**

.light[STA 379/679: Causal Inference <br>
Lucy D'Agostino McGowan
]

]

---

class: title title-1

# Weighted Estimator

.box-1[
$$\hat{\tau} = \frac{\sum_{i=1}^n Y_i X_i w_i}{\sum_{i=1}^nX_iw_i}-\frac{\sum_{i=1}^n Y_i (1-X_i) w_i}{\sum_{i=1}^n(1-X_i)w_i}$$
]

---
class: title title-1

# Weighted Estimator

.box-1[

$$\hat\mu_1 = \frac{\sum_{i=1}^n Y_i X_i w_i}{\sum_{i=1}^nX_iw_i}$$
]

--
<br>
.box-1[

$$\hat\mu_0 = \frac{\sum_{i=1}^n Y_i (1-X_i) w_i}{\sum_{i=1}^n(1-X_i)w_i}$$
]
---
class: title title-1

# Weighted Estimator (ATE)


$$\hat\mu_1 = \frac{\sum_{i=1}^n Y_i X_i \hat{e}_i^{-1}}{\sum_{i=1}^nX_i\hat{e}_i^{-1}}$$
]

--
<br>
.box-1[

$$\hat\mu_0 = \frac{\sum_{i=1}^n Y_i (1-X_i)(1-\hat{e}_i)^{-1}}{\sum_{i=1}^n(1-X_i)(1-\hat{e}_i)^{-1}}$$
]
---
class: title title-1
# Assuming the propensity score is *fixed*

.box-inv-1[
Equivalently, we could simultaneously solve the following estimating equations $\sum_{i=1}^n\mathbf{u}(\theta| Y_i, X_i, e_i) = 0$ for the parameter $\theta = (\mu_1, \mu_0)^T$ where
]

--
<br>
.box-1[
$$\mathbf{u}(\theta|Y, X, e)=\begin{pmatrix}(Y-\mu_1)Xe^{-1}\\(Y-\mu_0)(1-X)(1-e)^{-1}\end{pmatrix}$$
]

---

class: title title-1

# Assuming the propensity score is *fixed*

.box-1[
$\hat\theta$ is asymptotically normally distributed with a large sample variance equal to $n^{-1}\mathbf{A}^{-1}\mathbf{B}\mathbf{A}^{-T}$ where]

--

<br>

.box-1[
$$\mathbf{A}=-E\left[\frac{\partial\mathbf{u}}\right]$$
]






