---
title: "Missing Data"
author: "Lucy D'Agostino McGowan"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    css: ["default", "css/style.css"]
    seal: false
    anchor_sections: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```


```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view"))
library(tidyverse)
```

class: center middle main-title section-title-1

# Missing Data

.class-info[

**Session 18**

.light[STA 379/679: Causal Inference <br>
Lucy D'Agostino McGowan
]

]

---
class: title title-1

# Visualizing missingness

.left-code[
```{r, eval = FALSE}
library(visdat)
vis_dat(nhanes)
```
]

.right-plot[
```{r, echo = FALSE}
library(visdat)
library(mice)
vis_dat(nhanes)
```
]
---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

1. `install.packages("visdat")`
2. Load the causaldata package and use the `vis_dat()` function to examine the `nhefs` data frame
3. Fit a linear model predicting `pregnancies` from `birthcontrol`
4. Knit, commit, push to GitHub

`r countdown::countdown(5, font_size = "2em")`


---
class: title title-1

# Types of Missing Data

.pull-left-3[
.box-inv-1[Missing completely at random]

.box-1[The probability a variable is missing is the same for all observations]
]
--

.pull-middle-3[
.box-inv-1[Missing at random]

.box-1[The probability a variable is missing depends only on available information (the other variables)]
]

--

.pull-right-3[
.box-inv-1[Missing not at random]

.box-1[The probability a variable is missing depends on external factors]
]

---

class: title title-1

# Types of Missing Data

.pull-left-3[
.box-2[Missing completely at random]

.box-2[The probability a variable is missing is the same for all observations]
]


.pull-middle-3[
.box-inv-1[Missing at random]

.box-1[The probability a variable is missing depends only on available information (the other variables)]
]


.pull-right-3[
.box-2[Missing not at random]

.box-2[The probability a variable is missing depends on external factors]
]

---

class: title title-1

# Imputation

.box-1[Impute the mean/median or most common value for all missing variable]

--

.box-1[Use the observed variables to create a *model* to predict what the missing value would be]

--

.box-1[Use the observed variables to create many *models* to predict what the missing value would be (multiple imputation)]

---

class: title title-1

# Predictive mean matching

.box-inv-1[Variable A has some missing values]

--

.box-1[Calculates the predicted value of A for all observations using the other measured variables]

--

.box-1[For observation i, finds the set of observations that have the closest predicted value for variable A.]

--

.box-1[Randomly select one of the non-missing values of A from this set for the missing observation]

---

class: title title-1

# Example

.left-code[
```{r, eval = FALSE}
library(visdat)
vis_dat(nhanes)
```
]

.right-plot[
```{r, echo = FALSE}
library(visdat)
library(mice)
vis_dat(nhanes)
```
]

---
class:  title title-1

# Predictive mean matching

```{r, eval = FALSE}
library(mice)
set.seed(1)
nhanes_imp <- mice(nhanes, m = 1, method = "pmm")
```


---

class:  title title-1

# Predictive mean matching

```{r, eval = FALSE}
library(mice) #<<
set.seed(1)
nhanes_imp <- mice(nhanes, m = 1, method = "pmm")
```

---

class:  title title-1

# Predictive mean matching

```{r, output = "hide"}
library(mice)
set.seed(1)
nhanes_imp <- mice(nhanes, m = 1, method = "pmm") #<<
```

---

class:  title title-1

# Predictive mean matching

.pull-left[
```{r}
nhanes_imp$imp$bmi
```


]

.pull-right[
```{r}
nhanes %>%
  filter(bmi == 29.6)
```
]

---


class:  title title-1

# Predictive mean matching

```{r}
complete(nhanes_imp)
```
---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

1. `install.packages("mice")`
2. Use the `mice()` function to create a single imputed dataset of the `nhefs` data using predictive mean matching
3. Fit a linear model predicting `pregancies` from `birthcontrol` on the imputed dataset
4. Explore the help files for `?mice`. What types of univariate imputation methods are included?

`r countdown::countdown(8, font_size = "2em")`

---

class: title title-1

# Multiple imputation

```{r, eval = FALSE}
set.seed(1)
nhanes_imp <- mice(nhanes,
                   m = 5, 
                   method = "pmm")
```

---

class: title title-1

# Multiple imputation

```{r, eval = FALSE}
set.seed(1)
nhanes_imp <- mice(nhanes,
                   m = 5,  #<<
                   method = "pmm")
```
---

class: title title-1

# Multiple imputation

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
set.seed(1)
nhanes_imp <- mice(nhanes,
                   m = 5,  #<<
                   method = "pmm")
```

```{r, echo  = TRUE}
nhanes_imp$imp$bmi
```

---

class: title title-1

# Multiple imputation

```{r}
complete(nhanes_imp, 1)
```

---

class: title title-1

# Multiple imputation

```{r}
complete(nhanes_imp, 2)
```

---

class: title title-1

# Multiple imputation

```{r}
complete(nhanes_imp, 3)
```

---

class: title title-1

# Multiple imputation

```{r}
complete(nhanes_imp, 4)
```

---

class: title title-1

# Multiple imputation

```{r}
complete(nhanes_imp, 5)
```


---
class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

1. Use the `mice()` function to create a 5 imputed datasets of the `nhefs` data using predictive mean matching
3. Fit a linear model predicting `pregnancies` from `birthcontrol` on each of the 5 imputed datasets
4. Knit, commit, and push to GitHub

`r countdown::countdown(8, font_size = "2em")`

---

class: title title-1

# Pool Results

```{r}
fit <- with(data = nhanes_imp, lm(age ~ bmi))
```

---


class: title title-1

# Pool Results

```{r}
fit <- with(data = nhanes_imp, lm(age ~ bmi))
fit
```

---


class: title title-1

# Pool Results

```{r}
fit <- with(data = nhanes_imp, lm(age ~ bmi))
pool(fit)
```

---


class: title title-1

# Pool Results

```{r}
fit <- with(data = nhanes_imp, lm(age ~ bmi))
pool(fit) #<<
```

---


class: title title-1

# Pool Results

```{r}
with(data = nhanes_imp, lm(age ~ bmi)) %>%
  pool() %>%
  tidy()
```

---

class: title title-1

# Pooling effect estimates

.box-1.medium[
$$\bar\theta = \frac{1}{m}\sum_{i=1}^m\theta_i$$
]

---

class: title title-1

# Pooling standard errors

.box-1.medium[
$$V_{Total} = V_{within} + V_{between} + \frac{V_{between}}{m}$$
]

--

.box-1["Rubin's Rules"]

---

class: title title-1

# Pooling standard errors

.box-1.medium[
$$V_{within} =\frac{1}{m}\sum_{i=1}^mSE_i^2$$
]

---

class: title title-1

# Pooling standard errors

.box-1.medium[
$$V_{between} =\frac{\sum_{i=1}^m(\theta_i-\bar{\theta})^2}{m-1}$$
]

---

class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

1. Use the `with()` function to fit a model on each of your 5 imputed datasets.
2. Use the `pool()` function to pool the results. Compare to your previous results
3. Knit, commit, and push to GitHub

`r countdown::countdown(8, font_size = "2em")`

---

class: title title-1

# Adding in the PS weighting

.box-1[Across]

--

.box-inv-1[Create m imputed datasets, estimate the propensity score within each one then average across the m datasets, pool the results]

--

.box-1[Within]

--

.box-inv-1[Create m imputed datasets, estimate the propensity score within each one, run a seperate outcome model for each propensity score, pool the results.]

---


class: title title-1

# Adding in the PS weighting

```{r, message = FALSE, warning = FALSE, results = "hide"}
nhanes$exp <- rbinom(25, 1, 0.5)
nhanes_imp <- mice(nhanes, m = 5, method = "pmm")
```

---


class: title title-1

# MatchThem package

.box-1[From the makers of `MatchIt` we have a package that can handle the weights for you!]

```{r}
library(MatchThem)
nhanes_weighted <- weightthem(exp ~ bmi + hyp + chl, 
                              nhanes_imp,
                              approach = "across",
                              estimand = "ATE")
```

---

class: title title-1

# MatchThem package

.box-1[From the makers of `MatchIt` we have a package that can handle the weights for you!]

```{r}
library(MatchThem) #<<
nhanes_weighted <- weightthem(exp ~ bmi + hyp + chl, 
                              nhanes_imp,
                              approach = "across",
                              estimand = "ATE")
```

---

class: title title-1

# MatchThem package

.box-1[From the makers of `MatchIt` we have a package that can handle the weights for you!]

```{r}
library(MatchThem)
nhanes_weighted <- weightthem(exp ~ bmi + hyp + chl,  #<<
                              nhanes_imp,
                              approach = "across",
                              estimand = "ATE")
```

---

class: title title-1

# MatchThem package

.box-1[From the makers of `MatchIt` we have a package that can handle the weights for you!]

```{r}
library(MatchThem)
nhanes_weighted <- weightthem(exp ~ bmi + hyp + chl, 
                              nhanes_imp, #<<
                              approach = "across",
                              estimand = "ATE")
```

---
class: title title-1

# MatchThem package

.box-1[From the makers of `MatchIt` we have a package that can handle the weights for you!]

```{r}
library(MatchThem)
nhanes_weighted <- weightthem(exp ~ bmi + hyp + chl, 
                              nhanes_imp,
                              approach = "across", #<<
                              estimand = "ATE") 
```

---
class: title title-1

# MatchThem package

.box-1[From the makers of `MatchIt` we have a package that can handle the weights for you!]

```{r}
library(MatchThem)
nhanes_weighted <- weightthem(exp ~ bmi + hyp + chl, 
                              nhanes_imp,
                              approach = "across",
                              estimand = "ATE") #<<
```

---
class: title title-1

# MatchThem package


```{r}
library(survey)
with(nhanes_weighted, svyglm(age ~ exp)) %>%
       pool() %>%
       tidy()
```

---
class: title title-1

# MatchThem package


```{r}
library(survey) #<<
with(nhanes_weighted, svyglm(age ~ exp)) %>%
       pool() %>%
       tidy()
```

---

class: title title-1

# MatchThem package


```{r}
library(survey)
with(nhanes_weighted, svyglm(age ~ exp)) %>% #<<
       pool() %>%
       tidy()
```

---

class: title title-1

# MatchThem package


```{r}
library(survey)
with(nhanes_weighted, svyglm(age ~ exp)) %>%
       pool() %>% #<<
       tidy()
```

---

class: title title-1

# MatchThem package


```{r}
library(survey)
with(nhanes_weighted, svyglm(age ~ exp)) %>%
       pool() %>%
       tidy() #<<
```

---

class: title title-1

# MatchThem package

.box-1[From the makers of `MatchIt` we have a package that can handle the weights for you!]

```{r}
library(MatchThem)
nhanes_weighted <- weightthem(exp ~ bmi + hyp + chl, 
                              nhanes_imp,
                              approach = "within", #<<
                              estimand = "ATE") 
```

---
class: title title-1

# MatchThem package


```{r}
library(survey)
with(nhanes_weighted, svyglm(age ~ exp)) %>%
       pool() %>%
       tidy() 
```

---



class: title title-1

# `r fontawesome::fa("laptop")` Application Exercise

1. `install.packages("MatchThem")`
2. Use the `weightthem()` function to estimate the uncertainty for the causal effect of `birthcontrol` on `pregnancies`
2. Knit, Commit, Push to GitHub

`r countdown::countdown(8, font_size = "1.5em")`

---

